---
title: "Analysis of Buenrostro *et al* (2018) scATAC-seq data (original data, bulk peaks) using fastTopics with K = 10"
author: Kaixuan Luo
output:
  workflowr::wflow_html:
    toc: yes
---

Here we explore the structure in the Buenrostro *et al* (2018) 
scATAC-seq data inferred from the 
multinomial topic model with $k = 10$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load packages and some functions used in this analysis.

```{r load-pkgs, message=FALSE}
library(fastTopics)
library(Matrix)
library(dplyr)
library(ggplot2)
library(cowplot)
library(plyr)
library(dplyr)
library(RColorBrewer)
library(DT)
library(reshape)
source("code/plots.R")
```

### Load the data
Data downloaded from original paper.
```{r load-binarized-data}
data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/processed_data/"
load(file.path(data.dir, "Buenrostro_2018_binarized_counts.RData"))
cat(sprintf("%d x %d counts matrix.\n",nrow(counts),ncol(counts)))
samples$cell <- rownames(samples)
samples$label <- as.factor(samples$label)
```

### Topic model fit
Load the K = 10 topic model fit to the data downloaded from the original paper.
```{r load-fit}
fit.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Buenrostro_2018/binarized/"
fit <- readRDS(file.path(fit.dir, "/fit-Buenrostro2018-binarized-scd-ex-k=10.rds"))$fit
fit <- poisson2multinom(fit)
```

### Structure plot
```{r structure-plot, fig.width=7, fig.height=2.5, message=FALSE}
topic_colors <- c("darkorange","limegreen","magenta","gold","skyblue",
                  "darkblue","dodgerblue","darkmagenta","red","olivedrab")

set.seed(1)
# labels <- factor(samples$label, levels = c("HSC", "MPP", "CMP", "GMP", "mono", "MEP", "LMPP", "CLP", "pDC", "UNK"))

labels <- factor(samples$label, c("mono","pDC","MEP","HSC","MPP","CLP",
                                 "LMPP","CMP","GMP","UNK"))
structure_plot(fit,grouping = labels,colors = topic_colors,
               # topics = 1:10,
               gap = 20,perplexity = 50,verbose = FALSE)

```

### K-means clustering on topic proportions
Define clusters using k-means, and then create structure plot 
based on the clusters from k-means.

k-means clustering (using 12 clusters) on topic proportions
```{r structure-plot-kmeans-12clusters, fig.width=7, fig.height=2.5, message=FALSE}
set.seed(1)
clusters <- factor(kmeans(fit$L,centers = 12,iter.max = 100)$cluster)
summary(clusters)

structure_plot(fit,grouping = clusters,colors = topic_colors,
               gap = 20,perplexity = 50,verbose = FALSE)

```

### Differential accessibility (DA) analysis

DA analysis results **with "ash" shrinkage** (10000 MCMC iterations)

```{r load-DA-ash-res}
postfit_dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Buenrostro_2018/binarized/postfit_v2/"

DA_res <- readRDS(file.path(postfit_dir, "DAanalysis-Buenrostro2018-k=10/DA_regions_topics_ash_10000iters.rds"))
summary(DA_res)
```


Volcano plots for the regions

```{r region-volcano-plots-ash, fig.width=10, fig.height=6}
plots <- vector("list",10)
names(plots) <- 1:10

for (k in 1:10)
  plots[[k]] <- volcano_plot(DA_res, k, labels = rep("",nrow(DA_res$z)))
do.call(plot_grid,plots)

```

Number of regions selected at different lfsr cutoffs:
```{r number-DA-regions-ash}
sig_regions <- matrix(NA, nrow = 10, ncol = 5)
colnames(sig_regions) <- c("lfsr < 0.01", "lfsr < 0.05", "lfsr < 0.1", "lfsr < 0.2", "lfsr < 0.3")
rownames(sig_regions) <- paste("topic", 1:nrow(sig_regions))

for(k in 1:10){
  lfsr <- DA_res$lfsr[,k]
  sig_regions[k, ] <- c(length(which(lfsr < 0.01)), length(which(lfsr < 0.05)),
                         length(which(lfsr < 0.1)), length(which(lfsr < 0.2)), 
                         length(which(lfsr < 0.3)))
}

sig_regions
```


DA analysis results **without shrinkage** (10000 MCMC iterations)

```{r load-DA-no-shrinkage-res}
postfit_dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Buenrostro_2018/binarized/postfit_v2/"

DA_res <- readRDS(file.path(postfit_dir, "DAanalysis-Buenrostro2018-k=10/DA_regions_topics_noshrinkage_10000iters.rds"))
summary(DA_res)
```

Volcano plots for the regions
```{r region-volcano-plots-no-shrinkage, fig.width=10, fig.height=6}
plots <- vector("list",10)
names(plots) <- 1:10

for (k in 1:10)
  plots[[k]] <- volcano_plot(DA_res, k, labels = rep("",nrow(DA_res$z)))
do.call(plot_grid,plots)

```

Number of regions selected at different p-value cutoffs:
```{r number-DA-regions-no-shrinkage}
sig_regions <- matrix(NA, nrow = 10, ncol = 4)
colnames(sig_regions) <- c("p < 0.001", "p < 0.01", "p < 0.05", "p < 0.1")
rownames(sig_regions) <- paste("topic", 1:nrow(sig_regions))

for(k in 1:10){
  lpval <- DA_res$lpval[,k]
  pval <- 10^(-lpval)
  sig_regions[k, ] <- c(length(which(pval < 0.001)), length(which(pval < 0.01)),
                         length(which(pval < 0.05)), length(which(pval < 0.1)))
}

sig_regions
```
