---
title: "Assess Poisson NMF fits in Cusanovich_2018 scATAC-seq data"
author: "Kaixuan Luo"
output: workflowr::wflow_html
---

Here we compare the quality of the fits obtained from the different
updates (EM, CCD, SCD, with and without extrapolation), and with
different $k$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center")
```

## Assess Poisson NMF fits in Cusanovich_2018 scATAC-seq data in all tissues
Load the packages used in the analysis below, as well as some
additional functions for creating the plots.

```{r load-pkgs, message=FALSE}
library(fastTopics)
library(ggplot2)
library(cowplot)
source("code/functions_for_assessing_fits.R")
```

Load the results of running `fit_poisson_nmf` on the Cusanovich_2018 data,
with different algorithms, and for various choices of $k$ (the number
of "topics").

```{r load-results}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018"
load(file.path(out.dir, "/compiled.fits.Cusanovich2018.RData"))
```

This plot shows the improvement in the log-likelihood as the rank,
$k$, is increased. The log-likelihoods are shown relative to the
log-likelihood at $k = 2$. 

```{r plot-loglik-vs-k, fig.height=3, fig.width=4}
plot_loglik_vs_rank(fits) +
  theme_cowplot(font_size = 12)
```

As expected, the likelihood consistently improves as the rank, $k$, is
increased.

The next set of plots shows the improvement in the fit over time,
for all choices of $k$, and for each of the three optimization methods
(EM, CCD, SCD), with and without the extrapolation scheme. The quality
of the fit is measured by the log-likelihood relative to the best
log-likelihood that was identified among all methods compared.

```{r plot-loglik-iter, fig.height=8, fig.width=9}
create_progress_plots(dat, fits, x="iter", y="loglik", numiter.prefit=200)
```

The last set of plots shows the evolution of the KKT residuals over
time; the KKT residuals should vanish near a stationary point of the
log-likelihood, so looking at the largest KKT residual can be used to
assess how close we are to a solution. Note that, unlike the
log-likelihood, the KKT residuals are not expected to decrease
monotonically over time.

```{r plot-res-iter, fig.height=8, fig.width=9}
create_progress_plots(dat, fits, x="iter", y="res", numiter.prefit=200)
```

## Assess Poisson NMF fits in Cusanovich_2018 scATAC-seq data in bone marrow

```{r load-BoneMarrow-results}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018_BoneMarrow/"
load(file.path(out.dir, "/compiled.fits.Cusanovich2018.RData"))
```

Plot the improvement of log-likelihood as the rank, $k$, is increased. 

```{r plot-BoneMarrow-loglik-vs-k, fig.height=3, fig.width=4}
plot_loglik_vs_rank(fits) +
  theme_cowplot(font_size = 12)
```

Plots the improvement in the fit over time, for all choices of $k$.

```{r plot-BoneMarrow-loglik-iter, fig.height=8, fig.width=9}
create_progress_plots(dat, fits, x="iter", y="loglik", numiter.prefit=200)
```

Plots the evolution of the KKT residuals over time.
```{r plot-BoneMarrow-res-iter, fig.height=8, fig.width=9}
create_progress_plots(dat, fits, x="iter", y="res", numiter.prefit=200)
```


## Assess Poisson NMF fits in Cusanovich_2018 scATAC-seq data in whole brain

```{r load-WholeBrain-results}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018_WholeBrain/"
load(file.path(out.dir, "/compiled.fits.Cusanovich2018.RData"))
```

Plot the improvement of log-likelihood as the rank, $k$, is increased. 

```{r plot-WholeBrain-loglik-vs-k, fig.height=3, fig.width=4}
plot_loglik_vs_rank(fits) +
  theme_cowplot(font_size = 12)
```

Plots the improvement in the fit over time, for all choices of $k$.

```{r plot-WholeBrain-loglik-iter, fig.height=8, fig.width=9}
create_progress_plots(dat, fits, x="iter", y="loglik", numiter.prefit=200)
```

Plots the evolution of the KKT residuals over time.
```{r plot-WholeBrain-res-iter, fig.height=8, fig.width=9}
create_progress_plots(dat, fits, x="iter", y="res", numiter.prefit=200)
```



## Assess Poisson NMF fits in Cusanovich_2018 scATAC-seq data in heart, kindey and lung

```{r load-HeartKidneyLung-results}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018_HeartKidneyLung/"
load(file.path(out.dir, "/compiled.fits.Cusanovich2018.RData"))
```

Plot the improvement of log-likelihood as the rank, $k$, is increased. 

```{r plot-HeartKidneyLung-loglik-vs-k, fig.height=3, fig.width=4}
plot_loglik_vs_rank(fits) +
  theme_cowplot(font_size = 12)
```

Plots the improvement in the fit over time, for all choices of $k$.

```{r plot-HeartKidneyLung-loglik-iter, fig.height=8, fig.width=9}
create_progress_plots(dat, fits, x="iter", y="loglik", numiter.prefit=200)
```

Plots the evolution of the KKT residuals over time.
```{r plot-HeartKidneyLung-res-iter, fig.height=8, fig.width=9}
create_progress_plots(dat, fits, x="iter", y="res", numiter.prefit=200)
```
