---
title: "Topic modeling analysis of Buenrostro et al (2018) data with k = 10 topics"
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we summarize and interpret the results from the topic modeling
analysis of the Buenrostro et al (2018) data, with $k = 10$ topics.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages used in the analysis below.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(fastTopics)
library(ggplot2)
library(cowplot)
```

Load the count data (see [here](process_data_Buenrostro2018.html) for
details on these data and how they were prepared).

```{r load-data}
load("data/Buenrostro_2018/processed_data/Buenrostro_2018_binarized.RData")
```

Next we load the $k = 10$ multinomial topic model fit to these data,
the results of the GoM DE analysis using this topic model, and the
results of the HOMER motif enrichment analysis using the *p*-values
from the GoM DE analysis:

```{r load-results}
fit <- readRDS(
  file.path("output/Buenrostro_2018/binarized/filtered_peaks",
            "fit-Buenrostro2018-binarized-filtered-scd-ex-k=10.rds"))$fit
fit <- poisson2multinom(fit)
load(file.path("output/Buenrostro_2018/binarized/filtered_peaks",
               "de-buenrostro2018-k=10-noshrink.RData"))
homer <- readRDS(file.path("output/Buenrostro_2018/binarized/filtered_peaks",
                           "homer-buenrostro2018-k=10-noshrink.rds"))
```

Visualize the structure identified in the FACS cell populations using
a Structure plot:

```{r structure-plot, fig.height=1.75, fig.width=8, message=FALSE}
set.seed(1)
celltypes <- factor(samples$label,
                    c("pDC","CLP","LMPP","HSC","MPP","CMP","MEP","GMP",
                      "mono","UNK"))
topic_colors <- c("gold","forestgreen","orchid","red","lightgray",
                  "dimgray","orange","dodgerblue","limegreen","mediumblue")
p <- structure_plot(fit,grouping = celltypes,n = Inf,gap = 20,
                     perplexity = 70,verbose = FALSE,colors = topic_colors,
			         topics = c(5,10,8,3,4,1,6,7,2,9))
print(p)
```

```{r structure-plot-for-paper, echo=FALSE}
ggsave("plots/structure_plot_buenrostro.eps",p,height = 1.6,width = 8)
```

Comparing the topic proportions with the cell labels (provided by
FACS), we see that the topics capture chromatin accessibility patterns
characteristic of early hematopoietic progenitors (HSC, MPP; topics 2
and 9), and the four different hematopoietic cell lineages: CLPs and
LMPPs (topic 8); pDCs (topic 3); monocytes and GMPs (topics 1 and 7);
and MEPs (topic 4). Additional topics (topics 5, 6 and 10) may be
picking up other factors such as patient-specific batch effects or
tissue of origin (bone marrow or blood). Additionally, the split of
the HSC/MPP cells into two topics seems to also reflect
patient-specific batch effects, which also came up in the PCA analysis
in the original paper.

In the GoM DE analysis, we attempt to quantify *differential
accessibility* among topics. 

```{r de-analysis-plot, fig.height=2, fig.width=2.5, message=FALSE, warning=FALSE}
pdat <- data.frame(pval = 10^(-as.vector(de$lpval)))
p <- ggplot(pdat,aes(x = pval)) +
  geom_histogram(bins = 32,color = "white",fill = "black") +
  scale_y_continuous(breaks = seq(0,1e5,2e4)) +
  labs(x = "p-value",y = "LFCs") +
  theme_cowplot(font_size = 10)
print(p)
```

```{r de-analysis-plot-for-paper, echo=FALSE, warning=FALSE}
ggsave("plots/de_analysis_histogram_buenrostro.eps",p,height = 2.5,width = 3)
```

The issue is that individual regions rarely show significant
differences in accessibility, but a *motif enrichment analysis* of the
regions might point to interesting transcription factors common to the
more accessible regions. We performed the motif enrichment analysis
using HOMER.

First, we compile the motif enrichment *p*-values into a single table
(after removing a small number of duplicate results).

```{r compile-homer-results}
k <- 10
topics <- paste0("k",1:k)
motifs <- sort(unique(homer$k1[,"Motif Name"]))
n <- length(motifs)
homer_lpvals <- matrix(0,n,k)
rownames(homer_lpvals) <- motifs
colnames(homer_lpvals) <- topics
for (i in topics) {
  dat  <- homer[[i]]
  rows <- which(!duplicated(dat[,"Motif Name"]))
  dat  <- dat[rows,]
  rownames(dat)   <- dat[,"Motif Name"]
  homer_lpvals[,i] <- round(dat[motifs,"Log P-value"],digits = 2)
}
dat  <- homer$k1
rows <- which(!duplicated(dat[,"Motif Name"]))
dat  <- dat[rows,]
rownames(dat) <- dat[,"Motif Name"]
homer_lpvals <- cbind(dat[motifs,c("Motif Name","Consensus")],homer_lpvals)
```

```{r write-homer-results, echo=FALSE}
write.csv(homer_lpvals,"homer_buenrostro.csv",row.names = FALSE)
```
