---
title: "Compare PC plots of topic proportions from fastTopic with different K in Cusanovich et al (2018) data"
author: Kaixuan Luo
output:
  workflowr::wflow_html:
    toc: no
---

Here we explore the structure in the Cusanovich *et al* (2018)
ATAC-seq data inferred from the multinomial topic model with 
different numbers of $k$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```
					  
Load packages and some functions used in this analysis.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(ggplot2)
library(cowplot)
library(fastTopics)
source("code/plots.R")
```

Load the data. The counts are no longer needed at this stage of the
analysis.
```{r load-data}
data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Cusanovich_2018/processed_data/"
load(file.path(data.dir, "Cusanovich_2018.RData"))
rm(counts)
```

### k = 9
Load the $k = 9$ Poisson NMF fit.
```{r load-fit-k7}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018"
fit <- readRDS(file.path(out.dir, "/fit-Cusanovich2018-scd-ex-k=9.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-k9-1, fig.width=5, fig.height=4.5}
p1 <- basic_pca_plot(fit,1:2)
p2 <- basic_pca_plot(fit,3:4)
p3 <- basic_pca_plot(fit,5:6)
p4 <- basic_pca_plot(fit,7:8)
plot_grid(p1,p2,p3,p4)
```

These next plots visualize the same results, but as 2-d ("hexbin")
histograms:

```{r pca-k9-2, fig.width=5, fig.height=4.5}
p5 <- pca_hexbin_plot(fit,1:2) + guides(fill = "none")
p6 <- pca_hexbin_plot(fit,3:4) + guides(fill = "none")
p7 <- pca_hexbin_plot(fit,5:6) + guides(fill = "none")
p8 <- pca_hexbin_plot(fit,7:8) + guides(fill = "none")
plot_grid(p5,p6,p7,p8)
```


### k = 10
Load the $k = 10$ Poisson NMF fit.
```{r load-fit-k10}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018"
fit <- readRDS(file.path(out.dir, "/fit-Cusanovich2018-scd-ex-k=10.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-k10-1, fig.width=5, fig.height=4.5}
p1 <- basic_pca_plot(fit,1:2)
p2 <- basic_pca_plot(fit,3:4)
p3 <- basic_pca_plot(fit,5:6)
p4 <- basic_pca_plot(fit,7:8)
plot_grid(p1,p2,p3,p4)
```

These next plots visualize the same results, but as 2-d ("hexbin")
histograms:

```{r pca-k10-2, fig.width=5, fig.height=4.5}
p5 <- pca_hexbin_plot(fit,1:2) + guides(fill = "none")
p6 <- pca_hexbin_plot(fit,3:4) + guides(fill = "none")
p7 <- pca_hexbin_plot(fit,5:6) + guides(fill = "none")
p8 <- pca_hexbin_plot(fit,7:8) + guides(fill = "none")
plot_grid(p5,p6,p7,p8)
```

### k = 11
Load the $k = 11$ Poisson NMF fit.
```{r load-fit-k11}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018"
fit <- readRDS(file.path(out.dir, "/fit-Cusanovich2018-scd-ex-k=11.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-k11-1, fig.width=5, fig.height=4.5}
p1 <- basic_pca_plot(fit,1:2)
p2 <- basic_pca_plot(fit,3:4)
p3 <- basic_pca_plot(fit,5:6)
p4 <- basic_pca_plot(fit,7:8)
plot_grid(p1,p2,p3,p4)
```

These next plots visualize the same results, but as 2-d ("hexbin")
histograms:

```{r pca-k11-2, fig.width=5, fig.height=4.5}
p5 <- pca_hexbin_plot(fit,1:2) + guides(fill = "none")
p6 <- pca_hexbin_plot(fit,3:4) + guides(fill = "none")
p7 <- pca_hexbin_plot(fit,5:6) + guides(fill = "none")
p8 <- pca_hexbin_plot(fit,7:8) + guides(fill = "none")
plot_grid(p5,p6,p7,p8)
```


### k = 12
Load the $k = 12$ Poisson NMF fit.
```{r load-fit-k12}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018"
fit <- readRDS(file.path(out.dir, "/fit-Cusanovich2018-scd-ex-k=12.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-k12-1, fig.width=5, fig.height=4.5}
p1 <- basic_pca_plot(fit,1:2)
p2 <- basic_pca_plot(fit,3:4)
p3 <- basic_pca_plot(fit,5:6)
p4 <- basic_pca_plot(fit,7:8)
plot_grid(p1,p2,p3,p4)
```

These next plots visualize the same results, but as 2-d ("hexbin")
histograms:

```{r pca-k12-2, fig.width=5, fig.height=4.5}
p5 <- pca_hexbin_plot(fit,1:2) + guides(fill = "none")
p6 <- pca_hexbin_plot(fit,3:4) + guides(fill = "none")
p7 <- pca_hexbin_plot(fit,5:6) + guides(fill = "none")
p8 <- pca_hexbin_plot(fit,7:8) + guides(fill = "none")
plot_grid(p5,p6,p7,p8)
```


### k = 13
Load the $k = 13$ Poisson NMF fit.
```{r load-fit-k13}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018"
fit <- readRDS(file.path(out.dir, "/fit-Cusanovich2018-scd-ex-k=13.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-k13-1, fig.width=5, fig.height=4.5}
p1 <- basic_pca_plot(fit,1:2)
p2 <- basic_pca_plot(fit,3:4)
p3 <- basic_pca_plot(fit,5:6)
p4 <- basic_pca_plot(fit,7:8)
plot_grid(p1,p2,p3,p4)
```

These next plots visualize the same results, but as 2-d ("hexbin")
histograms:

```{r pca-k13-2, fig.width=5, fig.height=4.5}
p5 <- pca_hexbin_plot(fit,1:2) + guides(fill = "none")
p6 <- pca_hexbin_plot(fit,3:4) + guides(fill = "none")
p7 <- pca_hexbin_plot(fit,5:6) + guides(fill = "none")
p8 <- pca_hexbin_plot(fit,7:8) + guides(fill = "none")
plot_grid(p5,p6,p7,p8)
```


### k = 14
Load the $k = 14$ Poisson NMF fit.
```{r load-fit-k14}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018"
fit <- readRDS(file.path(out.dir, "/fit-Cusanovich2018-scd-ex-k=14.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-k14-1, fig.width=5, fig.height=4.5}
p1 <- basic_pca_plot(fit,1:2)
p2 <- basic_pca_plot(fit,3:4)
p3 <- basic_pca_plot(fit,5:6)
p4 <- basic_pca_plot(fit,7:8)
plot_grid(p1,p2,p3,p4)
```

These next plots visualize the same results, but as 2-d ("hexbin")
histograms:

```{r pca-k14-2, fig.width=5, fig.height=4.5}
p5 <- pca_hexbin_plot(fit,1:2) + guides(fill = "none")
p6 <- pca_hexbin_plot(fit,3:4) + guides(fill = "none")
p7 <- pca_hexbin_plot(fit,5:6) + guides(fill = "none")
p8 <- pca_hexbin_plot(fit,7:8) + guides(fill = "none")
plot_grid(p5,p6,p7,p8)
```


### k = 15
Load the $k = 15$ Poisson NMF fit.
```{r load-fit-k15}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018"
fit <- readRDS(file.path(out.dir, "/fit-Cusanovich2018-scd-ex-k=15.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-k15-1, fig.width=5, fig.height=4.5}
p1 <- basic_pca_plot(fit,1:2)
p2 <- basic_pca_plot(fit,3:4)
p3 <- basic_pca_plot(fit,5:6)
p4 <- basic_pca_plot(fit,7:8)
plot_grid(p1,p2,p3,p4)
```

These next plots visualize the same results, but as 2-d ("hexbin")
histograms:

```{r pca-k15-2, fig.width=5, fig.height=4.5}
p5 <- pca_hexbin_plot(fit,1:2) + guides(fill = "none")
p6 <- pca_hexbin_plot(fit,3:4) + guides(fill = "none")
p7 <- pca_hexbin_plot(fit,5:6) + guides(fill = "none")
p8 <- pca_hexbin_plot(fit,7:8) + guides(fill = "none")
plot_grid(p5,p6,p7,p8)
```
