---
title: "Clustering Cusanovich et al (2018) data using topic proportions with K = 13 topics"
author: Kaixuan Luo
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: false
---

Here we explore the structure in the Cusanovich *et al* (2018)
ATAC-seq data inferred from the multinomial topic model with 
$k = 13$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = FALSE,results = "hold",
                      fig.align = "center",dpi = 120, message = FALSE, warning = FALSE)
```
					  
Load packages and some functions used in this analysis.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(dplyr)
library(ggplot2)
library(cowplot)
library(fastTopics)
source("code/plots.R")
```

Load the data. The counts are no longer needed at this stage of the
analysis.
```{r load-data}
data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Cusanovich_2018/processed_data/"
load(file.path(data.dir, "Cusanovich_2018.RData"))
rm(counts)
```

Load the model fit
```{r load-fit}
fit.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018"
fit <- readRDS(file.path(fit.dir, "/fit-Cusanovich2018-scd-ex-k=13.rds"))$fit
fit_multinom <- poisson2multinom(fit)
```

About the samples:
The study measured single cell chromatin accessibility for 17 samples 
spanning 13 different tissues in 8-week old mice.

```{r samples}
cat(nrow(samples), "samples (cells). \n")
```

Tissues:
```{r samples-tissues}
samples$tissue <- as.factor(samples$tissue)
cat(length(levels(samples$tissue)), "tissues. \n")
# table(samples$tissue)

colors_tissues <- c("darkblue",    # BoneMarrow
                    "gray30",      # Cerebellum
                    "red",         # Heart
                    "springgreen", # Kidney
                    "brown",       # LargeIntestine
                    "purple",      # Liver
                    "deepskyblue", # Lung
                    "black",       # PreFrontalCortex
                    "darkgreen",   # SmallIntestine
                    "plum",        # Spleen
                    "gold",        # Testes
                    "darkorange",  # Thymus
                    "gray")        # WholeBrain
```

Cell types labels:
```{r samples-cells}
samples$cell_label <- as.factor(samples$cell_label)
cat(length(levels(samples$cell_label)), "cell types \n")
# table(samples$cell_label)

```

## Structure plots

The structure plots below summarize the topic proportions in the samples grouped by different tissues.

## Visualize by Structure plot grouped by tissues
```{r structure-plot-tissues, fig.width=8, fig.height=2.5, message=FALSE}
set.seed(10)
colors_topics <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
                   "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928",
                   "gray")
samples$tissue <- as.factor(samples$tissue)

rows <- sample(nrow(fit$L),4000)

p.structure.1 <- structure_plot(select(fit_multinom,loadings = rows),
                     grouping = samples[rows, "tissue"],n = Inf,gap = 40,
                     perplexity = 50,topics = 1:13,colors = colors_topics,
                     num_threads = 4,verbose = FALSE)

print(p.structure.1)
```

## Perform k-means clustering on topic proportions
Define clusters using k-means, and then create structure plot 
based on the clusters from k-means.

k-means clustering (using 15 clusters) on topic proportions
```{r structure-plot-kmeans-1, fig.width=8, fig.height=2.5, message=FALSE}

set.seed(10)
clusters <- factor(kmeans(fit_multinom$L,centers = 15,iter.max = 100)$cluster)
summary(clusters)

rows <- sample(nrow(fit$L),4000)

p.structure.kmeans <- structure_plot(select(fit_multinom,loadings = rows),
                                        grouping = clusters[rows],n = Inf,gap = 20,
                                        perplexity = 50,topics = 1:13,colors = colors_topics,
                                        num_threads = 4,verbose = FALSE)
print(p.structure.kmeans)
```

First do PCA on topic proportions and then do k-means clustering (using 15 clusters).
Results are the same as the results from running k-means directly on the topic proportions.
```{r structure-plot-kmeans-pca-1, fig.width=8, fig.height=2.5, message=FALSE}

set.seed(10)
pca <- prcomp(fit_multinom$L)$x
clusters2 <- factor(kmeans(pca,centers = 15,iter.max = 100)$cluster)
summary(clusters2)

rows <- sample(nrow(fit$L),4000)

p.structure.kmeans <- structure_plot(select(fit_multinom,loadings = rows),
                                        grouping = clusters2[rows],n = Inf,gap = 20,
                                        perplexity = 50,topics = 1:13,colors = colors_topics,
                                        num_threads = 4,verbose = FALSE)
print(p.structure.kmeans)

length(which(clusters != clusters2))
```

Refine k-means clusters:

  - Split cluster 10 into two clusters.
  - merge cluster 4 and 9.

Structure plot based on refined clusters
```{r structure-plot-kmeans-refined-1,  fig.width=8, fig.height=2.5, message=FALSE}
set.seed(10)
clusters.refined <- as.numeric(clusters)
clusters.refined[which(clusters == 9)] <- 4

# clusters.refined[which(clusters == 10 & fit_multinom$L[,8] >= 0.2)] <- 16
# clusters.refined[which(clusters == 10 & fit_multinom$L[,8] < 0.2)] <- 17

clusters.refined[which(clusters == 10)] <- kmeans(fit_multinom$L[which(clusters == 10), ],centers = 2)$cluster+15
clusters.refined <- factor(clusters.refined, labels = paste0("c", c(1:length(unique(clusters.refined)))))

samples$cluster_kmeans <- clusters.refined

rows <- sample(nrow(fit$L),4000)

p.structure.kmeans.refined <- structure_plot(select(fit_multinom,loadings = rows),
                                             grouping = clusters.refined[rows],n = Inf,gap = 40,
                                             perplexity = 50,topics = 1:13,colors = colors_topics,
                                             num_threads = 4,verbose = FALSE)
print(p.structure.kmeans.refined)

```

Save the clustering results to an RDS file
```{r, eval = FALSE}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018"
saveRDS(samples, paste0(out.dir, "/samples-clustering-Cusanovich2018.rds"))

samples <- readRDS("output/samples-clustering-Cusanovich2018.rds")
clusters.refined <- samples$cluster_kmeans
print(table(samples$cluster_kmeans))
```

## Plot the distribution of tissues in each cluster

Tissue composition in each cluster:
```{r freq-cluster-tissue, fig.width=8, fig.height=3}
library(plyr);library(dplyr)
library(RColorBrewer)

freq_cluster_tissue <- count(samples, vars=c("cluster_kmeans","tissue")) 

# stacked barplot for the counts of tissues by clusters
p.barplot <- ggplot(freq_cluster_tissue, aes(fill=tissue, y=freq, x=cluster_kmeans)) + 
  geom_bar(position="fill", stat="identity") +
  theme_classic() + xlab("Cluster") + ylab("Proportion of cells") +
  scale_fill_manual(values = colors_tissues) +
  guides(fill=guide_legend(ncol=2)) +
  theme(
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 8)
  )
print(p.barplot)

```

We can see a few clusters are tissue specific:

- cluster c3 is heart specific;
- cluster c8 is kidney specific;
- cluster c15 is liver specific;
- cluster c1 is primarily bone marrow;
- cluster c11 is primarily thymus.

Some clusters are combinations of related tissues:

- cluster c4 is half lung and half spleen;
- cluster c5 and c14 are mainly from pre-frontal cortex, whole brain (and cerebellum) -- all neuron related. 
- cluster c14 is also from whole brain and cerebellum.
- cluster c10 is mainly from Kidney, LargeIntestine, and Lung. 

Some clusters are more heterogeneous mixtures of different tissues.

Distribution of tissue labels by cluster.

```{r freq-table-cluster-tissue}
freq_table_cluster_tissue <- with(samples,table(tissue,cluster_kmeans))
print(freq_table_cluster_tissue)
```

Distribution of cell labels by cluster.

```{r freq-table-cluster-celllabel}
freq_table_cluster_celllabel <- with(samples,table(cell_label,cluster_kmeans))
print(freq_table_cluster_celllabel)
```


### zoom into a few clusters:

Structure plot for c2 cluster: 
Group samples by tissue labels first, then by cell labels
```{r structure-plot-c2-1, fig.width=8, fig.height=2.5}
rows.c2 <- which(samples$cluster_kmeans == "c2")

tissue_labels_cluster <- as.factor(as.character(samples[rows.c2, "tissue"]))
cell_labels_cluster <- as.factor(as.character(samples[rows.c2, "cell_label"]))

sort(table(tissue_labels_cluster), decreasing = T)
sort(table(cell_labels_cluster), decreasing = T)

p.structure <- structure_plot(select(fit_multinom,loadings = rows.c2),
                     grouping = tissue_labels_cluster,
                     rows = order(cell_labels_cluster),
                     n = Inf,gap = 40,
                     perplexity = 50,topics = 1:13,colors = colors_topics,
                     num_threads = 4,verbose = FALSE)
print(p.structure)

```

```{r pca-c2, fig.width=5, fig.height=2.25}
breaks <- c(0,1,5,10,100,Inf)

fit.c2 <- select(fit_multinom,loadings = rows.c2)
p.pca.c2 <- pca_plot(fit.c2,fill = "none")
p.pca.hexbin.c2 <- pca_hexbin_plot(fit.c2,breaks = breaks) + guides(fill = "none")
plot_grid(p.pca.c2, p.pca.hexbin.c2)
```

The variation in PCs 1 and 2 is mostly produced by topics 3,13
```{r pca-c2-topics, fig.width=6, fig.height=2.25}
p.pca.c2.topics <- pca_plot(fit.c2,k = c(3,13))
print(p.pca.c2.topics)
```

```{r pca-labels-c2, fig.width=8, fig.height=2.25}

colors_celllabels_c2 <- c("firebrick","dodgerblue","forestgreen",
                           "darkmagenta","darkorange","gold","darkblue",
                           "peru","greenyellow","olivedrab", 
                           "darkgray", "black")

p.pca.c2.tissue <- labeled_pca_plot(fit.c2,1:2,samples[rows.c2, "tissue"],font_size = 7,
                       colors = colors_tissues, legend_label = "tissue")
p.pca.c2.cell <- labeled_pca_plot(fit.c2,1:2,samples[rows.c2, "cell_label"],font_size = 7,
                        colors = colors_celllabels_c2, legend_label = "cell_label")
plot_grid(p.pca.c2.tissue,p.pca.c2.cell,rel_widths = c(8,9))
```

#### zoom into cluster c14

Structure plot of c14 cluster: 
Group samples by tissue labels first, then by cell labels
```{r structure-plot-c14-1, fig.width=7, fig.height=2.5}
rows.c14 <- which(samples$cluster_kmeans == "c14")

tissue_labels_cluster <- as.factor(as.character(samples[rows.c14, "tissue"]))
cell_labels_cluster <- as.factor(as.character(samples[rows.c14, "cell_label"]))

sort(table(tissue_labels_cluster), decreasing = T)
sort(table(cell_labels_cluster), decreasing = T)

p.structure <- structure_plot(select(fit_multinom,loadings = rows.c14),
                     grouping = tissue_labels_cluster,
                     rows = order(cell_labels_cluster),
                     n = Inf,gap = 40,
                     perplexity = 50,topics = 1:13,colors = colors_topics,
                     num_threads = 4,verbose = FALSE)
print(p.structure)


```

```{r pca-c14, fig.width=5, fig.height=2.25}
breaks <- c(0,1,5,10,100,Inf)

fit.c14 <- select(fit_multinom,loadings = rows.c14)
p.pca.c14 <- pca_plot(fit.c14,fill = "none")
p.pca.hexbin.c14 <- pca_hexbin_plot(fit.c14,breaks = breaks) + guides(fill = "none")
plot_grid(p.pca.c14, p.pca.hexbin.c14)
```

The variation in PCs 1 and 2 is mostly produced by topics 9,10,11
```{r pca-c14-topics, fig.width=6, fig.height=3}
p.pca.c14.topics <- pca_plot(fit.c14,k = c(9,10,11))
print(p.pca.c14.topics)
```

```{r pca-labels-c14, fig.width=8, fig.height=2.25}
colors_celllabels_c14 <- c("firebrick","dodgerblue","forestgreen",
                           "darkmagenta","darkorange","gold","darkblue",
                           "peru","greenyellow","black")

p.pca.c14.tissue <- labeled_pca_plot(fit.c14,1:2,samples[rows.c14, "tissue"],font_size = 7,
                       colors = colors_tissues, legend_label = "tissue")
p.pca.c14.cell <- labeled_pca_plot(fit.c14,1:2,samples[rows.c14, "cell_label"],font_size = 7,
                        colors = colors_celllabels_c14, legend_label = "cell_label")
plot_grid(p.pca.c14.tissue,p.pca.c14.cell,rel_widths = c(8,11))

```

```{r}

print(samples[intersect(rows.c14, which(samples$tissue == "Lung")),])

```

## PCA plots

Plot PCs of the topic proportions.
```{r pca-k13-1, fig.width=7, fig.height=4.5}
p.pca1.1 <- pca_plot(fit_multinom,pcs = 1:2,fill = "none")
p.pca1.2 <- pca_plot(fit_multinom,pcs = 3:4,fill = "none")
p.pca1.3 <- pca_plot(fit_multinom,pcs = 5:6,fill = "none")
p.pca1.4 <- pca_plot(fit_multinom,pcs = 7:8,fill = "none")
p.pca1.5 <- pca_plot(fit_multinom,pcs = 9:10,fill = "none")
p.pca1.6 <- pca_plot(fit_multinom,pcs = 11:12,fill = "none")

plot_grid(p.pca1.1,p.pca1.2,p.pca1.3,p.pca1.4,p.pca1.5,p.pca1.6)
```

Some of the structure is more evident from “hexbin” plots 
showing the density of the points.

```{r pca-k13-2, fig.width=7, fig.height=4.5}
breaks <- c(0,1,5,10,100,Inf)
p.pca2.1 <- pca_hexbin_plot(fit_multinom, pcs = 1:2, breaks = breaks) + guides(fill = "none")
p.pca2.2 <- pca_hexbin_plot(fit_multinom, pcs = 3:4, breaks = breaks) + guides(fill = "none")
p.pca2.3 <- pca_hexbin_plot(fit_multinom, pcs = 5:6, breaks = breaks) + guides(fill = "none")
p.pca2.4 <- pca_hexbin_plot(fit_multinom, pcs = 7:8, breaks = breaks) + guides(fill = "none")
p.pca2.5 <- pca_hexbin_plot(fit_multinom, pcs = 9:10, breaks = breaks) + guides(fill = "none")
p.pca2.6 <- pca_hexbin_plot(fit_multinom, pcs = 11:12, breaks = breaks) + guides(fill = "none")

plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4,p.pca2.5,p.pca2.6)
```

## Layer the tissue and cell labels onto the PC plots

PCs 1 and 2:

```{r pca-labels-k13-1, fig.width=8, fig.height=2.25}
p.pca3.1 <- labeled_pca_plot(fit,1:2,samples$tissue,font_size = 7,
                       colors = colors_tissues, legend_label = "tissue")
p.pca3.2 <- labeled_pca_plot(fit,1:2,samples$cell_label,font_size = 7,
                        legend_label = "cell_label")
plot_grid(p.pca3.1,p.pca3.2,rel_widths = c(8,11))
```

PCs 3 and 4:

```{r pca-labels-k13-2, fig.width=8, fig.height=2.25}
p.pca3.3 <- labeled_pca_plot(fit,3:4,samples$tissue,font_size = 7,
                        colors = colors_tissues, legend_label = "tissue")
p.pca3.4 <- labeled_pca_plot(fit,3:4,samples$cell_label,font_size = 7,
                        legend_label = "cell_label")
plot_grid(p.pca3.3,p.pca3.4,rel_widths = c(8,11))
```

PCs 5 and 6:

```{r pca-labels-k13-3, fig.width=8, fig.height=2.25}
p.pca3.5 <- labeled_pca_plot(fit,5:6,samples$tissue,font_size = 7,
                       colors = colors_tissues, legend_label = "tissue")
p.pca3.6 <- labeled_pca_plot(fit,5:6,samples$cell_label,font_size = 7,
                        legend_label = "cell_label")
plot_grid(p.pca3.5,p.pca3.6,rel_widths = c(8,11))
```

PCs 7 and 8:

```{r pca-labels-k13-4, fig.width=8, fig.height=2.25}
p.pca3.7 <- labeled_pca_plot(fit,7:8,samples$tissue,font_size = 7,
                       colors = colors_tissues, legend_label = "tissue")
p.pca3.8 <- labeled_pca_plot(fit,7:8,samples$cell_label,font_size = 7,
                        legend_label = "cell_label")
plot_grid(p.pca3.7,p.pca3.8,rel_widths = c(8,11))
```

PCs 9 and 10:

```{r pca-labels-k13-5, fig.width=8, fig.height=2.25}
p.pca3.9 <- labeled_pca_plot(fit,9:10,samples$tissue,font_size = 7,
                       colors = colors_tissues, legend_label = "tissue")
p.pca3.10 <- labeled_pca_plot(fit,9:10,samples$cell_label,font_size = 7,
                        legend_label = "cell_label")
plot_grid(p.pca3.9,p.pca3.10,rel_widths = c(8,11))
```


PCs 11 and 12:

```{r pca-labels-k13-6, fig.width=8, fig.height=2.25}
p.pca3.11 <- labeled_pca_plot(fit,11:12,samples$tissue,font_size = 7,
                       legend_label = "tissue")
p.pca3.12 <- labeled_pca_plot(fit,11:12,samples$cell_label,font_size = 7,
                        legend_label = "cell_label")
plot_grid(p.pca3.11,p.pca3.12,rel_widths = c(8,11))
```

The clusters defined by k-means on topic proportions
reasonably identify the clusters shown in the PCA hexbin plots (below).

```{r pca-kmeans-1, fig.width=9, fig.height=4.5}
p.pca.4.1 <- labeled_pca_plot(fit,1:2,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.2 <- labeled_pca_plot(fit,3:4,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.3 <- labeled_pca_plot(fit,5:6,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.4 <- labeled_pca_plot(fit,7:8,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.5 <- labeled_pca_plot(fit,9:10,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.6 <- labeled_pca_plot(fit,11:12,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
plot_grid(p.pca.4.1,p.pca.4.2,p.pca.4.3,p.pca.4.4,p.pca.4.5,p.pca.4.6)
```
