---
title: "Clustering Cusanovich et al (2018) data using topic proportions with K = 13"
author: Kaixuan Luo
output:
  workflowr::wflow_html:
    toc: no
---

Here we explore the structure in the Cusanovich *et al* (2018)
ATAC-seq data inferred from the multinomial topic model with 
$k = 13$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```
					  
Load packages and some functions used in this analysis.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(dplyr)
library(ggplot2)
library(cowplot)
library(fastTopics)
source("code/plots.R")
```

Load the data. The counts are no longer needed at this stage of the
analysis.
```{r load-data}
data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Cusanovich_2018/processed_data/"
load(file.path(data.dir, "Cusanovich_2018.RData"))
rm(counts)
```

## Plot PCs of the topic proportions
We first use PCA to explore the structure inferred from the
multinomial topic model with $k = 13$:

Load the $k = 13$ Poisson NMF fit.
```{r load-fit-k13}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018"
fit <- readRDS(file.path(out.dir, "/fit-Cusanovich2018-scd-ex-k=13.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-k13-1, fig.width=7, fig.height=4.5}
p.pca1.1 <- pca_plot(poisson2multinom(fit),pcs = 1:2,fill = "none")
p.pca1.2 <- pca_plot(poisson2multinom(fit),pcs = 3:4,fill = "none")
p.pca1.3 <- pca_plot(poisson2multinom(fit),pcs = 5:6,fill = "none")
p.pca1.4 <- pca_plot(poisson2multinom(fit),pcs = 7:8,fill = "none")
p.pca1.5 <- pca_plot(poisson2multinom(fit),pcs = 9:10,fill = "none")
p.pca1.6 <- pca_plot(poisson2multinom(fit),pcs = 11:12,fill = "none")

plot_grid(p.pca1.1,p.pca1.2,p.pca1.3,p.pca1.4,p.pca1.5,p.pca1.6)
```

Some of the structure is more evident from “hexbin” plots 
showing the density of the points.

```{r pca-k13-2, fig.width=7, fig.height=4.5}
breaks <- c(0,1,5,10,100,Inf)
p.pca2.1 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 1:2, breaks = breaks) + guides(fill = "none")
p.pca2.2 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 3:4, breaks = breaks) + guides(fill = "none")
p.pca2.3 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 5:6, breaks = breaks) + guides(fill = "none")
p.pca2.4 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 7:8, breaks = breaks) + guides(fill = "none")
p.pca2.5 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 9:10, breaks = breaks) + guides(fill = "none")
p.pca2.6 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 11:12, breaks = breaks) + guides(fill = "none")

plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4,p.pca2.5,p.pca2.6)
```

## Layer the tissue and cell labels onto the PC plots
Next, we layer the tissue and cell labels onto the PC plots.

PCs 1 and 2:

```{r pca-labels-k13-1, fig.width=8, fig.height=2.25}
p.pca3.1 <- labeled_pca_plot(fit,1:2,samples$tissue,font_size = 7,
                       legend_label = "tissue")
p.pca3.2 <- labeled_pca_plot(fit,1:2,samples$cell_label,font_size = 7,
                        legend_label = "cell_label")
plot_grid(p.pca3.1,p.pca3.2,rel_widths = c(8,11))
```

PCs 3 and 4:

```{r pca-labels-k13-2, fig.width=8, fig.height=2.25}
p.pca3.3 <- labeled_pca_plot(fit,3:4,samples$tissue,font_size = 7,
                        legend_label = "tissue")
p.pca3.4 <- labeled_pca_plot(fit,3:4,samples$cell_label,font_size = 7,
                        legend_label = "cell_label")
plot_grid(p.pca3.3,p.pca3.4,rel_widths = c(8,11))
```

PCs 5 and 6:

```{r pca-labels-k13-3, fig.width=8, fig.height=2.25}
p.pca3.5 <- labeled_pca_plot(fit,5:6,samples$tissue,font_size = 7,
                       legend_label = "tissue")
p.pca3.6 <- labeled_pca_plot(fit,5:6,samples$cell_label,font_size = 7,
                        legend_label = "cell_label")
plot_grid(p.pca3.5,p.pca3.6,rel_widths = c(8,11))
```

PCs 7 and 8:

```{r pca-labels-k13-4, fig.width=8, fig.height=2.25}
p.pca3.7 <- labeled_pca_plot(fit,7:8,samples$tissue,font_size = 7,
                       legend_label = "tissue")
p.pca3.8 <- labeled_pca_plot(fit,7:8,samples$cell_label,font_size = 7,
                        legend_label = "cell_label")
plot_grid(p.pca3.7,p.pca3.8,rel_widths = c(8,11))
```

PCs 9 and 10:

```{r pca-labels-k13-5, fig.width=8, fig.height=2.25}
p.pca3.9 <- labeled_pca_plot(fit,9:10,samples$tissue,font_size = 7,
                       legend_label = "tissue")
p.pca3.10 <- labeled_pca_plot(fit,9:10,samples$cell_label,font_size = 7,
                        legend_label = "cell_label")
plot_grid(p.pca3.9,p.pca3.10,rel_widths = c(8,11))
```


PCs 11 and 12:

```{r pca-labels-k13-6, fig.width=8, fig.height=2.25}
p.pca3.11 <- labeled_pca_plot(fit,11:12,samples$tissue,font_size = 7,
                       legend_label = "tissue")
p.pca3.12 <- labeled_pca_plot(fit,11:12,samples$cell_label,font_size = 7,
                        legend_label = "cell_label")
plot_grid(p.pca3.11,p.pca3.12,rel_widths = c(8,11))
```

Visualize by structure plot grouped by tissues
```{r structure-plot-1, fig.width=7, fig.height=2.5}
set.seed(10)
colors_topics <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
                   "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928",
                   "gray")
rows <- sample(nrow(fit$L),4000)
samples$tissue <- as.factor(samples$tissue)

p.structure.1 <- structure_plot(select(poisson2multinom(fit),loadings = rows),
                     grouping = samples[rows, "tissue"],n = Inf,gap = 20,
                     perplexity = 50,topics = 1:13,colors = colors_topics,
                     num_threads = 4,verbose = FALSE)

print(p.structure.1)
```

## k-means clustering on topic proportions
Define clusters using k-means, and then create structure plot 
based on the clusters from k-means.

Define clusters using k-means with $k = 14$:
```{r kmeans-1}
set.seed(10)

clusters.14 <- factor(kmeans(poisson2multinom(fit)$L,centers = 14)$cluster)
print(sort(table(clusters.14),decreasing = TRUE))
```

Structure plot based on k-means clusters
```{r structure-plot-kmeans-1, fig.width=7, fig.height=2}
colors_topics <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
                   "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928",
                   "gray")
rows <- sample(nrow(fit$L),4000)

p.structure.2 <- structure_plot(select(poisson2multinom(fit),loadings = rows),
                     grouping = clusters.14[rows],n = Inf,gap = 20,
                     perplexity = 50,topics = 1:13,colors = colors_topics,
                     num_threads = 4,verbose = FALSE)
print(p.structure.2)
```

Define clusters using k-means with $k = 15$:
```{r kmeans-2}
set.seed(10)

clusters.15 <- factor(kmeans(poisson2multinom(fit)$L,centers = 15)$cluster)
print(sort(table(clusters.15),decreasing = TRUE))

```

Structure plot based on k-means clusters
```{r structure-plot-kmeans-2, fig.width=7, fig.height=2}
colors_topics <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
            "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928",
            "gray")
rows <- sample(nrow(fit$L),4000)

p.structure.3 <- structure_plot(select(poisson2multinom(fit),loadings = rows),
                     grouping = clusters.15[rows],n = Inf,gap = 20,
                     perplexity = 50,topics = 1:13,colors = colors_topics,
                     num_threads = 4,verbose = FALSE)
print(p.structure.3)
```

Define clusters using k-means with $k = 20$:
```{r kmeans-3}
set.seed(10)

clusters.20 <- factor(kmeans(poisson2multinom(fit)$L,centers = 20)$cluster)
print(sort(table(clusters.20),decreasing = TRUE))
```

Structure plot based on k-means clusters
```{r structure-plot-kmeans-3, fig.width=7, fig.height=2}
colors_topics <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
                   "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928",
                   "gray")
rows <- sample(nrow(fit$L),4000)
p.structure.4 <- structure_plot(select(poisson2multinom(fit),loadings = rows),
                     grouping = clusters.20[rows],n = Inf,gap = 20,
                     perplexity = 50,topics = 1:13,colors = colors_topics,
                     num_threads = 4,verbose = FALSE)
print(p.structure.4)
```

## Refine clusters
We then further refine the clusters based on k-means result with $k = 20$:
merge "orange" clusters 9, 11, 14;
merge "brown" clusters 3 and 10, 16, 19;
merge "yellow" clusters 8 and 18.
(maybe could also merge the "red" clusters 2 and 4)
```{r refine-cluster-1}
clusters.merged <- clusters.20
clusters.merged[clusters.20 %in% c(9,11,14)] <- 9
clusters.merged[clusters.20 %in% c(3,10,16,19)] <- 3
clusters.merged[clusters.20 %in% c(8,18)] <- 8
clusters.merged <- factor(clusters.merged, labels = paste0("c", c(1:length(unique(clusters.merged)))))
print(sort(table(clusters.merged),decreasing = TRUE))

samples$cluster_kmeans <- clusters.merged

```

```{r refine-cluster-2, fig.width=7, fig.height=2}
colors_topics <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
                   "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928",
                   "gray")
rows <- sample(nrow(fit$L),4000)
p.structure.5 <- structure_plot(select(poisson2multinom(fit),loadings = rows),
                     grouping = clusters.merged[rows],n = Inf,gap = 20,
                     perplexity = 50,topics = 1:13,colors = colors_topics,
                     num_threads = 4,verbose = FALSE)
print(p.structure.5)
```

The clusters defined by k-means on topic proportions
reasonably identify the clusters shown in the PCA hexbin plots (below).

```{r pca-kmeans-1, fig.width=9, fig.height=4.5}
p.pca.4.1 <- labeled_pca_plot(fit,1:2,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.2 <- labeled_pca_plot(fit,3:4,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.3 <- labeled_pca_plot(fit,5:6,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.4 <- labeled_pca_plot(fit,7:8,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.5 <- labeled_pca_plot(fit,9:10,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.6 <- labeled_pca_plot(fit,11:12,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
plot_grid(p.pca.4.1,p.pca.4.2,p.pca.4.3,p.pca.4.4,p.pca.4.5,p.pca.4.6)
```

```{r pca-kmeans-2, fig.width=7, fig.height=4.5}
plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4,p.pca2.5,p.pca2.6)

```

We then label the cells in each cluster with the known tissue labels.

Tissues:
```{r samples-tissues}
samples$tissue <- as.factor(samples$tissue)
cat(length(levels(samples$tissue)), "tissues. \n")
table(samples$tissue)

```

Plot the distribution of tissues by cluster.

Stacked barplot for the counts of tissues by clusters:
```{r freq-cluster-tissue-1, fig.width=8, fig.height=3}
library(plyr);library(dplyr)
library(RColorBrewer)

freq_cluster_tissue <- count(samples, vars=c("cluster_kmeans","tissue")) 

colors_tissues <- colorRampPalette(brewer.pal(9, "Set1"))(13)

# stacked barplot for the counts of tissues by clusters
p.structure.6 <- ggplot(freq_cluster_tissue, aes(fill=tissue, y=freq, x=cluster_kmeans)) + 
  geom_bar(position="stack", stat="identity") +
  theme_classic() + xlab("Cluster") + ylab("Number of cells") +
  scale_fill_manual(values = colors_tissues) +
  guides(fill=guide_legend(ncol=2)) +
  theme(
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 8)
  )
print(p.structure.6)

```

Percent stacked barplot for the counts of tissues by clusters:
```{r freq-cluster-tissue-2, fig.width=8, fig.height=3}

freq_cluster_tissue <- count(samples, vars=c("cluster_kmeans","tissue")) 

colors_tissues <- colorRampPalette(brewer.pal(9, "Set1"))(13)

p.structure.7 <- ggplot(freq_cluster_tissue, aes(fill=tissue, y=freq, x=cluster_kmeans)) + 
  geom_bar(position="fill", stat="identity") +
  theme_classic() + xlab("Cluster") + ylab("Proportion of cells") +
  scale_fill_manual(values = colors_tissues) +
  guides(fill=guide_legend(ncol=2)) +
  theme(
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 8)
  )
print(p.structure.7)

```

We can see a few clusters are tissue specific:
cluster c5 is kidney specific;
cluster c7 is heart specific;
cluster c9 is liver specific;
cluster c3 is primarily thymus;
cluster c6 is primarily bone marrow.

Some clusters are combinations of related tissues:
cluster c4 is half lung and half spleen;
cluster c8 and c12 are mainly from pre-frontal cortex, whole brain (and cerebellum) -- all neuron related. 
cluster c10 is also from whole brain and cerebellum.
cluster c14 is mainly from Kidney, LargeIntestine, and Lung. 

Some clusters are more heterogeneous mixtures of different tissues:
e.g. c1, c2, c11, c13. 

```{r freq-cluster-tissue}
freq_cluster_tissue <- with(samples,table(tissue,cluster_kmeans))
print(freq_cluster_tissue)
```

