---
title: "Clustering Lareau *et al* (2019) bone marrow data using topic proportions with K = 13"
author: Kaixuan Luo
output:
  workflowr::wflow_html:
    toc: no
---

Here we explore the structure in the Lareau *et al* (2019) 
human bone marrow scATAC-seq data inferred from the 
multinomial topic model with $k = 13$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```
					  
Load packages and some functions used in this analysis.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(dplyr)
library(ggplot2)
library(cowplot)
library(fastTopics)
source("code/plots.R")
```

The study used dsciATAC-seq to profile bone marrow mononuclear cells (BMMCs)
from two human donors before (*resting*, untreated controls) and after stimulation (*stimulated*).

## *resting* condition

### Load the data in *resting* condition
```{r load-data-resting}
data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Lareau_2019/bone_marrow/processed_data/"
load(file.path(data.dir, "Lareau_2019_bonemarrow_resting.RData"))
cat(sprintf("%d x %d counts matrix.\n",nrow(counts),ncol(counts)))
rm(counts)
samples$Cluster <- as.factor(samples$Cluster)
```

### Plot PCs of the topic proportions
We first use PCA to explore the structure inferred from the
multinomial topic model with $k = 13$:

```{r load-fit-resting}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Lareau_2019/Lareau_2019_bonemarrow_resting"
fit <- readRDS(file.path(out.dir, "/fit-Lareau2019_bonemarrow_resting-scd-ex-k=13.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-resting-1, fig.width=7, fig.height=4.5}
p.pca1.1 <- pca_plot(poisson2multinom(fit),pcs = 1:2,fill = "none")
p.pca1.2 <- pca_plot(poisson2multinom(fit),pcs = 3:4,fill = "none")
p.pca1.3 <- pca_plot(poisson2multinom(fit),pcs = 5:6,fill = "none")
p.pca1.4 <- pca_plot(poisson2multinom(fit),pcs = 7:8,fill = "none")
p.pca1.5 <- pca_plot(poisson2multinom(fit),pcs = 9:10,fill = "none")
p.pca1.6 <- pca_plot(poisson2multinom(fit),pcs = 11:12,fill = "none")

plot_grid(p.pca1.1,p.pca1.2,p.pca1.3,p.pca1.4,p.pca1.5,p.pca1.6)
```

Some of the structure is more evident from “hexbin” plots 
showing the density of the points.

```{r pca-resting-2, fig.width=7, fig.height=4.5}
breaks <- c(0,1,5,10,100,Inf)
p.pca2.1 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 1:2, breaks = breaks) + guides(fill = "none")
p.pca2.2 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 3:4, breaks = breaks) + guides(fill = "none")
p.pca2.3 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 5:6, breaks = breaks) + guides(fill = "none")
p.pca2.4 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 7:8, breaks = breaks) + guides(fill = "none")
p.pca2.5 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 9:10, breaks = breaks) + guides(fill = "none")
p.pca2.6 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 11:12, breaks = breaks) + guides(fill = "none")

plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4,p.pca2.5,p.pca2.6)
```

### Layer the cell labels onto the PC plots
Next, we layer the cell labels onto the PC plots.

```{r pca-resting-3, fig.width=9, fig.height=5}
p.pca3.1 <- labeled_pca_plot(fit,1:2,samples$Cluster,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.2 <- labeled_pca_plot(fit,3:4,samples$Cluster,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.3 <- labeled_pca_plot(fit,5:6,samples$Cluster,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.4 <- labeled_pca_plot(fit,7:8,samples$Cluster,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.5 <- labeled_pca_plot(fit,9:10,samples$Cluster,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.6 <- labeled_pca_plot(fit,11:12,samples$Cluster,font_size = 7,
                       legend_label = "Cell labels")
plot_grid(p.pca3.1,p.pca3.2,p.pca3.3,p.pca3.4,p.pca3.5,p.pca3.6)
```

Visualize by structure plot grouped by cell labels
```{r structure-plot-resting-1, fig.width=7, fig.height=3}
set.seed(10)
colors_topics <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
                   "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928",
                   "gray")
rows <- sample(nrow(fit$L),4000)
samples$Cluster <- as.factor(samples$Cluster)

p.structure.1 <- structure_plot(select(poisson2multinom(fit),loadings = rows),
                     grouping = samples[rows, "Cluster"],n = Inf,gap = 20,
                     perplexity = 50,topics = 1:13,colors = colors_topics,
                     num_threads = 4,verbose = FALSE)

print(p.structure.1)
```

### k-means clustering on topic proportions
Define clusters using k-means, and then create structure plot 
based on the clusters from k-means.

Define clusters using k-means with $k = 10$:
```{r kmeans-resting-1}
set.seed(10)

clusters.10 <- factor(kmeans(poisson2multinom(fit)$L,centers = 10)$cluster)
print(sort(table(clusters.10),decreasing = TRUE))
```

Structure plot based on k-means clusters
```{r structure-plot-kmeans-resting-1, fig.width=7, fig.height=2}
colors_topics <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
                   "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928",
                   "gray")
rows <- sample(nrow(fit$L),4000)

p.structure.2 <- structure_plot(select(poisson2multinom(fit),loadings = rows),
                     grouping = clusters.10[rows],n = Inf,gap = 20,
                     perplexity = 50,topics = 1:13,colors = colors_topics,
                     num_threads = 4,verbose = FALSE)
print(p.structure.2)
```

Define clusters using k-means with $k = 12$:
```{r kmeans-resting-2}
set.seed(10)

clusters.12 <- factor(kmeans(poisson2multinom(fit)$L,centers = 12)$cluster)
print(sort(table(clusters.12),decreasing = TRUE))
```

Structure plot based on k-means clusters
```{r structure-plot-kmeans-resting-2, fig.width=7, fig.height=2}
colors_topics <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
                   "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928",
                   "gray")
rows <- sample(nrow(fit$L),4000)

p.structure.3 <- structure_plot(select(poisson2multinom(fit),loadings = rows),
                     grouping = clusters.12[rows],n = Inf,gap = 20,
                     perplexity = 50,topics = 1:13,colors = colors_topics,
                     num_threads = 4,verbose = FALSE)
print(p.structure.3)
```

### Refine clusters
We then further refine the clusters based on k-means result with $k = 12$:
merge clusters 4 and 6;
(maybe also merge clusters 7 and 8).
```{r refine-cluster-resting-1}
clusters.merged <- clusters.12
clusters.merged[clusters.12 %in% c(4,6)] <- 4
# clusters.merged[clusters.12 %in% c(7,8)] <- 7
clusters.merged <- factor(clusters.merged, labels = paste0("c", c(1:length(unique(clusters.merged)))))
print(sort(table(clusters.merged),decreasing = TRUE))

samples$cluster_kmeans <- clusters.merged

```

```{r refine-cluster-resting-2, fig.width=7, fig.height=2}
colors_topics <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
                   "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928",
                   "gray")
rows <- sample(nrow(fit$L),4000)
p.structure.refined <- structure_plot(select(poisson2multinom(fit),loadings = rows),
                     grouping = clusters.merged[rows],n = Inf,gap = 20,
                     perplexity = 50,topics = 1:13,colors = colors_topics,
                     num_threads = 4,verbose = FALSE)
print(p.structure.refined)
```

The clusters defined by k-means on topic proportions
reasonably identify the clusters shown in the PCA hexbin plots (below).

```{r pca-kmeans-resting-1, fig.width=9, fig.height=4.5}
p.pca.4.1 <- labeled_pca_plot(fit,1:2,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.2 <- labeled_pca_plot(fit,3:4,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.3 <- labeled_pca_plot(fit,5:6,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.4 <- labeled_pca_plot(fit,7:8,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.5 <- labeled_pca_plot(fit,9:10,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.6 <- labeled_pca_plot(fit,11:12,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
plot_grid(p.pca.4.1,p.pca.4.2,p.pca.4.3,p.pca.4.4,p.pca.4.5,p.pca.4.6)
```

```{r pca-kmeans-resting-2, fig.width=7, fig.height=4.5}
plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4,p.pca2.5,p.pca2.6)

```

We then label the cells in each cluster with the known cell labels.

Cell labels:
```{r samples-cells-resting}
samples$Cluster <- as.factor(samples$Cluster)
cat(length(levels(samples$Cluster)), "cell labels. \n")
table(samples$Cluster)

```

Plot the distribution of cell labels by cluster.

Stacked barplot for the counts of tissues by cell labels:
```{r freq-cluster-cell-resting-1, fig.width=8, fig.height=3}
library(plyr);library(dplyr)
library(RColorBrewer)

freq_cluster_tissue <- count(samples, vars=c("cluster_kmeans","Cluster")) 
n_colors <- length(levels(samples$Cluster))
colors_labels <- colorRampPalette(brewer.pal(9, "Set1"))(n_colors)

# stacked barplot for the counts of tissues by clusters
p.barplot.1 <- ggplot(freq_cluster_tissue, aes(fill=Cluster, y=freq, x=cluster_kmeans)) + 
  geom_bar(position="stack", stat="identity") +
  theme_classic() + xlab("Cluster") + ylab("Number of cells") +
  scale_fill_manual(values = colors_labels) +
  guides(fill=guide_legend(ncol=2)) +
  theme(
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 8)
  )
print(p.barplot.1)

```

Percent stacked barplot for the counts of tissues by clusters:
```{r freq-cluster-tissue-resting-2, fig.width=8, fig.height=3}

freq_cluster_tissue <- count(samples, vars=c("cluster_kmeans","Cluster")) 
n_colors <- length(levels(samples$Cluster))
colors_labels <- colorRampPalette(brewer.pal(9, "Set1"))(n_colors)

p.barplot.2 <- ggplot(freq_cluster_tissue, aes(fill=Cluster, y=freq, x=cluster_kmeans)) + 
  geom_bar(position="fill", stat="identity") +
  theme_classic() + xlab("Cluster") + ylab("Proportion of cells") +
  scale_fill_manual(values = colors_labels) +
  guides(fill=guide_legend(ncol=2)) +
  theme(
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 8)
  )
print(p.barplot.2)

```

```{r freq-cluster-cells-resting}
freq_cluster_cells <- with(samples,table(Cluster,cluster_kmeans))
print(freq_cluster_cells)
```



## *stimulated* condition

### Load the data in *stimulated* condition
```{r load-data-stimulated}
data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Lareau_2019/bone_marrow/processed_data/"
load(file.path(data.dir, "Lareau_2019_bonemarrow_stimulated.RData"))
cat(sprintf("%d x %d counts matrix.\n",nrow(counts),ncol(counts)))
rm(counts)
samples$Cluster <- as.factor(samples$Cluster)
```

### Plot PCs of the topic proportions
We first use PCA to explore the structure inferred from the
multinomial topic model with $k = 13$:

```{r load-fit-stimulated}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Lareau_2019/Lareau_2019_bonemarrow_stimulated"
fit <- readRDS(file.path(out.dir, "/fit-Lareau2019_bonemarrow_stimulated-scd-ex-k=13.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-stimulated-1, fig.width=7, fig.height=4.5}
p.pca1.1 <- pca_plot(poisson2multinom(fit),pcs = 1:2,fill = "none")
p.pca1.2 <- pca_plot(poisson2multinom(fit),pcs = 3:4,fill = "none")
p.pca1.3 <- pca_plot(poisson2multinom(fit),pcs = 5:6,fill = "none")
p.pca1.4 <- pca_plot(poisson2multinom(fit),pcs = 7:8,fill = "none")
p.pca1.5 <- pca_plot(poisson2multinom(fit),pcs = 9:10,fill = "none")
p.pca1.6 <- pca_plot(poisson2multinom(fit),pcs = 11:12,fill = "none")

plot_grid(p.pca1.1,p.pca1.2,p.pca1.3,p.pca1.4,p.pca1.5,p.pca1.6)
```

Some of the structure is more evident from “hexbin” plots 
showing the density of the points.

```{r pca-stimulated-2, fig.width=7, fig.height=4.5}
breaks <- c(0,1,5,10,100,Inf)
p.pca2.1 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 1:2, breaks = breaks) + guides(fill = "none")
p.pca2.2 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 3:4, breaks = breaks) + guides(fill = "none")
p.pca2.3 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 5:6, breaks = breaks) + guides(fill = "none")
p.pca2.4 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 7:8, breaks = breaks) + guides(fill = "none")
p.pca2.5 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 9:10, breaks = breaks) + guides(fill = "none")
p.pca2.6 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 11:12, breaks = breaks) + guides(fill = "none")

plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4,p.pca2.5,p.pca2.6)
```

### Layer the cell labels onto the PC plots
Next, we layer the cell labels onto the PC plots.

```{r pca-stimulated-3, fig.width=9, fig.height=5}
p.pca3.1 <- labeled_pca_plot(fit,1:2,samples$Cluster,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.2 <- labeled_pca_plot(fit,3:4,samples$Cluster,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.3 <- labeled_pca_plot(fit,5:6,samples$Cluster,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.4 <- labeled_pca_plot(fit,7:8,samples$Cluster,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.5 <- labeled_pca_plot(fit,9:10,samples$Cluster,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.6 <- labeled_pca_plot(fit,11:12,samples$Cluster,font_size = 7,
                       legend_label = "Cell labels")
plot_grid(p.pca3.1,p.pca3.2,p.pca3.3,p.pca3.4,p.pca3.5,p.pca3.6)
```

Visualize by structure plot grouped by cell labels
```{r structure-plot-stimulated-1, fig.width=7, fig.height=3}
set.seed(10)
colors_topics <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
                   "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928",
                   "gray")
rows <- sample(nrow(fit$L),4000)
samples$Cluster <- as.factor(samples$Cluster)

p.structure.1 <- structure_plot(select(poisson2multinom(fit),loadings = rows),
                     grouping = samples[rows, "Cluster"],n = Inf,gap = 20,
                     perplexity = 50,topics = 1:13,colors = colors_topics,
                     num_threads = 4,verbose = FALSE)

print(p.structure.1)
```

### k-means clustering on topic proportions
Define clusters using k-means, and then create structure plot 
based on the clusters from k-means.

Define clusters using k-means with $k = 10$:
```{r kmeans-stimulated-1}
set.seed(10)

clusters.10 <- factor(kmeans(poisson2multinom(fit)$L,centers = 10)$cluster)
print(sort(table(clusters.10),decreasing = TRUE))
```

Structure plot based on k-means clusters
```{r structure-plot-kmeans-stimulated-1, fig.width=7, fig.height=2}
colors_topics <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
                   "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928",
                   "gray")
rows <- sample(nrow(fit$L),4000)

p.structure.2 <- structure_plot(select(poisson2multinom(fit),loadings = rows),
                     grouping = clusters.10[rows],n = Inf,gap = 20,
                     perplexity = 50,topics = 1:13,colors = colors_topics,
                     num_threads = 4,verbose = FALSE)
print(p.structure.2)
```

Define clusters using k-means with $k = 12$:
```{r kmeans-stimulated-2}
set.seed(10)

clusters.12 <- factor(kmeans(poisson2multinom(fit)$L,centers = 12)$cluster)
print(sort(table(clusters.12),decreasing = TRUE))
```

Structure plot based on k-means clusters
```{r structure-plot-kmeans-stimulated-2, fig.width=7, fig.height=2}
colors_topics <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
                   "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928",
                   "gray")
rows <- sample(nrow(fit$L),4000)

p.structure.3 <- structure_plot(select(poisson2multinom(fit),loadings = rows),
                     grouping = clusters.12[rows],n = Inf,gap = 20,
                     perplexity = 50,topics = 1:13,colors = colors_topics,
                     num_threads = 4,verbose = FALSE)
print(p.structure.3)
```

### Refine clusters
We then further refine the clusters based on k-means result with $k = 12$:
merge clusters 5 and 8.
```{r refine-cluster-stimulated-1}
clusters.merged <- clusters.12
clusters.merged[clusters.12 %in% c(5,8)] <- 5
clusters.merged <- factor(clusters.merged, labels = paste0("c", c(1:length(unique(clusters.merged)))))
print(sort(table(clusters.merged),decreasing = TRUE))

samples$cluster_kmeans <- clusters.merged

```

```{r refine-cluster-stimulated-2, fig.width=7, fig.height=2}
colors_topics <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
                   "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928",
                   "gray")
rows <- sample(nrow(fit$L),4000)
p.structure.refined <- structure_plot(select(poisson2multinom(fit),loadings = rows),
                     grouping = clusters.merged[rows],n = Inf,gap = 20,
                     perplexity = 50,topics = 1:13,colors = colors_topics,
                     num_threads = 4,verbose = FALSE)
print(p.structure.refined)
```

The clusters defined by k-means on topic proportions
reasonably identify the clusters shown in the PCA hexbin plots (below).

```{r pca-kmeans-stimulated-1, fig.width=9, fig.height=4.5}
p.pca.4.1 <- labeled_pca_plot(fit,1:2,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.2 <- labeled_pca_plot(fit,3:4,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.3 <- labeled_pca_plot(fit,5:6,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.4 <- labeled_pca_plot(fit,7:8,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.5 <- labeled_pca_plot(fit,9:10,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
p.pca.4.6 <- labeled_pca_plot(fit,11:12,samples$cluster_kmeans,font_size = 7,
                       legend_label = "cluster_kmeans")
plot_grid(p.pca.4.1,p.pca.4.2,p.pca.4.3,p.pca.4.4,p.pca.4.5,p.pca.4.6)
```

```{r pca-kmeans-stimulated-2, fig.width=7, fig.height=4.5}
plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4,p.pca2.5,p.pca2.6)

```

We then label the cells in each cluster with the known cell labels.

Cell labels:
```{r samples-cells-stimulated}
samples$Cluster <- as.factor(samples$Cluster)
cat(length(levels(samples$Cluster)), "cell labels. \n")
table(samples$Cluster)

```

Plot the distribution of cell labels by cluster.

Stacked barplot for the counts of tissues by cell labels:
```{r freq-cluster-cell-stimulated-1, fig.width=8, fig.height=3}
library(plyr);library(dplyr)
library(RColorBrewer)

freq_cluster_tissue <- count(samples, vars=c("cluster_kmeans","Cluster")) 
n_colors <- length(levels(samples$Cluster))
colors_labels <- colorRampPalette(brewer.pal(9, "Set1"))(n_colors)

# stacked barplot for the counts of tissues by clusters
p.barplot.1 <- ggplot(freq_cluster_tissue, aes(fill=Cluster, y=freq, x=cluster_kmeans)) + 
  geom_bar(position="stack", stat="identity") +
  theme_classic() + xlab("Cluster") + ylab("Number of cells") +
  scale_fill_manual(values = colors_labels) +
  guides(fill=guide_legend(ncol=2)) +
  theme(
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 8)
  )
print(p.barplot.1)

```

Percent stacked barplot for the counts of tissues by clusters:
```{r freq-cluster-tissue-stimulated-2, fig.width=8, fig.height=3}

freq_cluster_tissue <- count(samples, vars=c("cluster_kmeans","Cluster")) 
n_colors <- length(levels(samples$Cluster))
colors_labels <- colorRampPalette(brewer.pal(9, "Set1"))(n_colors)

p.barplot.2 <- ggplot(freq_cluster_tissue, aes(fill=Cluster, y=freq, x=cluster_kmeans)) + 
  geom_bar(position="fill", stat="identity") +
  theme_classic() + xlab("Cluster") + ylab("Proportion of cells") +
  scale_fill_manual(values = colors_labels) +
  guides(fill=guide_legend(ncol=2)) +
  theme(
  legend.title = element_text(size = 10),
  legend.text = element_text(size = 8)
  )
print(p.barplot.2)

```

```{r freq-cluster-cells-stimulated}
freq_cluster_cells <- with(samples,table(Cluster,cluster_kmeans))
print(freq_cluster_cells)
```

