---
title: "Compare PC plots of topic proportions from fastTopic with different K in Buenrostro *et al* (2018) scATAC-seq data"
author: Kaixuan Luo
output:
  workflowr::wflow_html:
    toc: yes
---

Here we explore the structure in the Buenrostro *et al* (2018) 
scATAC-seq data inferred from the multinomial topic model with 
different numbers of $k$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```
					  
Load packages and some functions used in this analysis.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(ggplot2)
library(cowplot)
library(fastTopics)
source("code/plots.R")
```

We applied the analysis on binarized and unbinarized scATAC-seq data.
The data was processed using chromVAR on aggregated single-cell ATAC-seq peaks.

## Binarized counts
Load the data
```{r load-binarized-data}
data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/processed_data_Chen2019pipeline/chromVAR/"
load(file.path(data.dir, "Buenrostro_2018_binarized_scPeaks.RData"))
cat(sprintf("%d x %d counts matrix.\n",nrow(counts),ncol(counts)))
rm(counts)
samples$label <- as.factor(samples$label)
```

### Plot PCs of the topic proportions
We first use PCA to explore the structure inferred from the
multinomial topic model with different numbers of $k$:

#### k = 10
Load the $k = 10$ Poisson NMF fit.
```{r load-binarized-fit-k10}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Buenrostro_2018_chromVAR_scPeaks/binarized/"
fit <- readRDS(file.path(out.dir, "/fit-Buenrostro2018-binarized-scd-ex-k=10.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-binarized-k10-1, fig.width=7, fig.height=4.5}
p.pca1.1 <- pca_plot(poisson2multinom(fit),pcs = 1:2,fill = "none")
p.pca1.2 <- pca_plot(poisson2multinom(fit),pcs = 3:4,fill = "none")
p.pca1.3 <- pca_plot(poisson2multinom(fit),pcs = 5:6,fill = "none")
p.pca1.4 <- pca_plot(poisson2multinom(fit),pcs = 7:8,fill = "none")

plot_grid(p.pca1.1,p.pca1.2,p.pca1.3,p.pca1.4)
```

Some of the structure is more evident from “hexbin” plots 
showing the density of the points.

```{r pca-binarized-k10-2, fig.width=7, fig.height=4.5}
breaks <- c(0,1,5,10,100,Inf)
p.pca2.1 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 1:2, breaks = breaks) + guides(fill = "none")
p.pca2.2 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 3:4, breaks = breaks) + guides(fill = "none")
p.pca2.3 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 5:6, breaks = breaks) + guides(fill = "none")
p.pca2.4 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 7:8, breaks = breaks) + guides(fill = "none")

plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4)
```

Next, we layer the cell labels onto the PC plots.

```{r pca-binarized-k10-3, fig.width=8, fig.height=6}
p.pca3.1 <- labeled_pca_plot(fit,1:2,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.2 <- labeled_pca_plot(fit,3:4,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.3 <- labeled_pca_plot(fit,5:6,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.4 <- labeled_pca_plot(fit,7:8,samples$label,font_size = 7,
                       legend_label = "Cell labels")

plot_grid(p.pca3.1,p.pca3.2,p.pca3.3,p.pca3.4)
```



#### k = 11
Load the $k = 11$ Poisson NMF fit.
```{r load-binarized-fit-k11}
fit <- readRDS(file.path(out.dir, "/fit-Buenrostro2018-binarized-scd-ex-k=11.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-binarized-k11-1, fig.width=7, fig.height=4.5}
p.pca1.1 <- pca_plot(poisson2multinom(fit),pcs = 1:2,fill = "none")
p.pca1.2 <- pca_plot(poisson2multinom(fit),pcs = 3:4,fill = "none")
p.pca1.3 <- pca_plot(poisson2multinom(fit),pcs = 5:6,fill = "none")
p.pca1.4 <- pca_plot(poisson2multinom(fit),pcs = 7:8,fill = "none")

plot_grid(p.pca1.1,p.pca1.2,p.pca1.3,p.pca1.4)
```

Some of the structure is more evident from “hexbin” plots 
showing the density of the points.

```{r pca-binarized-k11-2, fig.width=7, fig.height=4.5}
breaks <- c(0,1,5,10,100,Inf)
p.pca2.1 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 1:2, breaks = breaks) + guides(fill = "none")
p.pca2.2 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 3:4, breaks = breaks) + guides(fill = "none")
p.pca2.3 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 5:6, breaks = breaks) + guides(fill = "none")
p.pca2.4 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 7:8, breaks = breaks) + guides(fill = "none")

plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4)
```

Next, we layer the cell labels onto the PC plots.

```{r pca-binarized-k11-3, fig.width=8, fig.height=6}
p.pca3.1 <- labeled_pca_plot(fit,1:2,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.2 <- labeled_pca_plot(fit,3:4,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.3 <- labeled_pca_plot(fit,5:6,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.4 <- labeled_pca_plot(fit,7:8,samples$label,font_size = 7,
                       legend_label = "Cell labels")

plot_grid(p.pca3.1,p.pca3.2,p.pca3.3,p.pca3.4)
```

#### k = 12
Load the $k = 12$ Poisson NMF fit.
```{r load-binarized-fit-k12}
fit <- readRDS(file.path(out.dir, "/fit-Buenrostro2018-binarized-scd-ex-k=12.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-binarized-k12-1, fig.width=7, fig.height=4.5}
p.pca1.1 <- pca_plot(poisson2multinom(fit),pcs = 1:2,fill = "none")
p.pca1.2 <- pca_plot(poisson2multinom(fit),pcs = 3:4,fill = "none")
p.pca1.3 <- pca_plot(poisson2multinom(fit),pcs = 5:6,fill = "none")
p.pca1.4 <- pca_plot(poisson2multinom(fit),pcs = 7:8,fill = "none")

plot_grid(p.pca1.1,p.pca1.2,p.pca1.3,p.pca1.4)
```

Some of the structure is more evident from “hexbin” plots 
showing the density of the points.

```{r pca-binarized-k12-2, fig.width=7, fig.height=4.5}
breaks <- c(0,1,5,10,100,Inf)
p.pca2.1 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 1:2, breaks = breaks) + guides(fill = "none")
p.pca2.2 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 3:4, breaks = breaks) + guides(fill = "none")
p.pca2.3 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 5:6, breaks = breaks) + guides(fill = "none")
p.pca2.4 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 7:8, breaks = breaks) + guides(fill = "none")

plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4)
```

Next, we layer the cell labels onto the PC plots.

```{r pca-binarized-k12-3, fig.width=8, fig.height=6}
p.pca3.1 <- labeled_pca_plot(fit,1:2,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.2 <- labeled_pca_plot(fit,3:4,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.3 <- labeled_pca_plot(fit,5:6,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.4 <- labeled_pca_plot(fit,7:8,samples$label,font_size = 7,
                       legend_label = "Cell labels")

plot_grid(p.pca3.1,p.pca3.2,p.pca3.3,p.pca3.4)
```

#### k = 13
Load the $k = 13$ Poisson NMF fit.
```{r load-binarized-fit-k13}
fit <- readRDS(file.path(out.dir, "/fit-Buenrostro2018-binarized-scd-ex-k=13.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-binarized-k13-1, fig.width=7, fig.height=4.5}
p.pca1.1 <- pca_plot(poisson2multinom(fit),pcs = 1:2,fill = "none")
p.pca1.2 <- pca_plot(poisson2multinom(fit),pcs = 3:4,fill = "none")
p.pca1.3 <- pca_plot(poisson2multinom(fit),pcs = 5:6,fill = "none")
p.pca1.4 <- pca_plot(poisson2multinom(fit),pcs = 7:8,fill = "none")

plot_grid(p.pca1.1,p.pca1.2,p.pca1.3,p.pca1.4)
```

Some of the structure is more evident from “hexbin” plots 
showing the density of the points.

```{r pca-binarized-k13-2, fig.width=7, fig.height=4.5}
breaks <- c(0,1,5,10,100,Inf)
p.pca2.1 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 1:2, breaks = breaks) + guides(fill = "none")
p.pca2.2 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 3:4, breaks = breaks) + guides(fill = "none")
p.pca2.3 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 5:6, breaks = breaks) + guides(fill = "none")
p.pca2.4 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 7:8, breaks = breaks) + guides(fill = "none")

plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4)
```

Next, we layer the cell labels onto the PC plots.

```{r pca-binarized-k13-3, fig.width=8, fig.height=6}
p.pca3.1 <- labeled_pca_plot(fit,1:2,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.2 <- labeled_pca_plot(fit,3:4,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.3 <- labeled_pca_plot(fit,5:6,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.4 <- labeled_pca_plot(fit,7:8,samples$label,font_size = 7,
                       legend_label = "Cell labels")

plot_grid(p.pca3.1,p.pca3.2,p.pca3.3,p.pca3.4)
```


#### k = 14
Load the $k = 14$ Poisson NMF fit.
```{r load-binarized-fit-k14}
fit <- readRDS(file.path(out.dir, "/fit-Buenrostro2018-binarized-scd-ex-k=14.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-binarized-k14-1, fig.width=7, fig.height=4.5}
p.pca1.1 <- pca_plot(poisson2multinom(fit),pcs = 1:2,fill = "none")
p.pca1.2 <- pca_plot(poisson2multinom(fit),pcs = 3:4,fill = "none")
p.pca1.3 <- pca_plot(poisson2multinom(fit),pcs = 5:6,fill = "none")
p.pca1.4 <- pca_plot(poisson2multinom(fit),pcs = 7:8,fill = "none")

plot_grid(p.pca1.1,p.pca1.2,p.pca1.3,p.pca1.4)
```

Some of the structure is more evident from “hexbin” plots 
showing the density of the points.

```{r pca-binarized-k14-2, fig.width=7, fig.height=4.5}
breaks <- c(0,1,5,10,100,Inf)
p.pca2.1 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 1:2, breaks = breaks) + guides(fill = "none")
p.pca2.2 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 3:4, breaks = breaks) + guides(fill = "none")
p.pca2.3 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 5:6, breaks = breaks) + guides(fill = "none")
p.pca2.4 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 7:8, breaks = breaks) + guides(fill = "none")

plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4)
```

Next, we layer the cell labels onto the PC plots.

```{r pca-binarized-k14-3, fig.width=8, fig.height=6}
p.pca3.1 <- labeled_pca_plot(fit,1:2,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.2 <- labeled_pca_plot(fit,3:4,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.3 <- labeled_pca_plot(fit,5:6,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.4 <- labeled_pca_plot(fit,7:8,samples$label,font_size = 7,
                       legend_label = "Cell labels")

plot_grid(p.pca3.1,p.pca3.2,p.pca3.3,p.pca3.4)
```


#### k = 15
Load the $k = 15$ Poisson NMF fit.
```{r load-binarized-fit-k15}
fit <- readRDS(file.path(out.dir, "/fit-Buenrostro2018-binarized-scd-ex-k=15.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-binarized-k15-1, fig.width=7, fig.height=4.5}
p.pca1.1 <- pca_plot(poisson2multinom(fit),pcs = 1:2,fill = "none")
p.pca1.2 <- pca_plot(poisson2multinom(fit),pcs = 3:4,fill = "none")
p.pca1.3 <- pca_plot(poisson2multinom(fit),pcs = 5:6,fill = "none")
p.pca1.4 <- pca_plot(poisson2multinom(fit),pcs = 7:8,fill = "none")

plot_grid(p.pca1.1,p.pca1.2,p.pca1.3,p.pca1.4)
```

Some of the structure is more evident from “hexbin” plots 
showing the density of the points.

```{r pca-binarized-k15-2, fig.width=7, fig.height=4.5}
breaks <- c(0,1,5,10,100,Inf)
p.pca2.1 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 1:2, breaks = breaks) + guides(fill = "none")
p.pca2.2 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 3:4, breaks = breaks) + guides(fill = "none")
p.pca2.3 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 5:6, breaks = breaks) + guides(fill = "none")
p.pca2.4 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 7:8, breaks = breaks) + guides(fill = "none")

plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4)
```

Next, we layer the cell labels onto the PC plots.

```{r pca-binarized-k15-3, fig.width=8, fig.height=6}
p.pca3.1 <- labeled_pca_plot(fit,1:2,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.2 <- labeled_pca_plot(fit,3:4,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.3 <- labeled_pca_plot(fit,5:6,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.4 <- labeled_pca_plot(fit,7:8,samples$label,font_size = 7,
                       legend_label = "Cell labels")

plot_grid(p.pca3.1,p.pca3.2,p.pca3.3,p.pca3.4)
```


## Unbinarized counts
Load the data
```{r load-unbinarized-data}
data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/processed_data_Chen2019pipeline/chromVAR/"
load(file.path(data.dir, "Buenrostro_2018_scPeaks.RData"))
cat(sprintf("%d x %d counts matrix.\n",nrow(counts),ncol(counts)))
rm(counts)
samples$label <- as.factor(samples$label)
```

### Plot PCs of the topic proportions
We first use PCA to explore the structure inferred from the
multinomial topic model with different numbers of $k$:

#### k = 10
Load the $k = 10$ Poisson NMF fit.
```{r load-unbinarized-fit-k10}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Buenrostro_2018_chromVAR_scPeaks/unbinarized/"
fit <- readRDS(file.path(out.dir, "/fit-Buenrostro2018-scd-ex-k=10.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-unbinarized-k10-1, fig.width=7, fig.height=4.5}
p.pca1.1 <- pca_plot(poisson2multinom(fit),pcs = 1:2,fill = "none")
p.pca1.2 <- pca_plot(poisson2multinom(fit),pcs = 3:4,fill = "none")
p.pca1.3 <- pca_plot(poisson2multinom(fit),pcs = 5:6,fill = "none")
p.pca1.4 <- pca_plot(poisson2multinom(fit),pcs = 7:8,fill = "none")

plot_grid(p.pca1.1,p.pca1.2,p.pca1.3,p.pca1.4)
```

Some of the structure is more evident from “hexbin” plots 
showing the density of the points.

```{r pca-unbinarized-k10-2, fig.width=7, fig.height=4.5}
breaks <- c(0,1,5,10,100,Inf)
p.pca2.1 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 1:2, breaks = breaks) + guides(fill = "none")
p.pca2.2 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 3:4, breaks = breaks) + guides(fill = "none")
p.pca2.3 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 5:6, breaks = breaks) + guides(fill = "none")
p.pca2.4 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 7:8, breaks = breaks) + guides(fill = "none")

plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4)
```

Next, we layer the cell labels onto the PC plots.

```{r pca-unbinarized-k10-3, fig.width=8, fig.height=6}
p.pca3.1 <- labeled_pca_plot(fit,1:2,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.2 <- labeled_pca_plot(fit,3:4,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.3 <- labeled_pca_plot(fit,5:6,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.4 <- labeled_pca_plot(fit,7:8,samples$label,font_size = 7,
                       legend_label = "Cell labels")

plot_grid(p.pca3.1,p.pca3.2,p.pca3.3,p.pca3.4)
```



#### k = 11
Load the $k = 11$ Poisson NMF fit.
```{r load-unbinarized-fit-k11}
fit <- readRDS(file.path(out.dir, "/fit-Buenrostro2018-scd-ex-k=11.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-unbinarized-k11-1, fig.width=7, fig.height=4.5}
p.pca1.1 <- pca_plot(poisson2multinom(fit),pcs = 1:2,fill = "none")
p.pca1.2 <- pca_plot(poisson2multinom(fit),pcs = 3:4,fill = "none")
p.pca1.3 <- pca_plot(poisson2multinom(fit),pcs = 5:6,fill = "none")
p.pca1.4 <- pca_plot(poisson2multinom(fit),pcs = 7:8,fill = "none")

plot_grid(p.pca1.1,p.pca1.2,p.pca1.3,p.pca1.4)
```

Some of the structure is more evident from “hexbin” plots 
showing the density of the points.

```{r pca-unbinarized-k11-2, fig.width=7, fig.height=4.5}
breaks <- c(0,1,5,10,100,Inf)
p.pca2.1 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 1:2, breaks = breaks) + guides(fill = "none")
p.pca2.2 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 3:4, breaks = breaks) + guides(fill = "none")
p.pca2.3 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 5:6, breaks = breaks) + guides(fill = "none")
p.pca2.4 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 7:8, breaks = breaks) + guides(fill = "none")

plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4)
```

Next, we layer the cell labels onto the PC plots.

```{r pca-unbinarized-k11-3, fig.width=8, fig.height=6}
p.pca3.1 <- labeled_pca_plot(fit,1:2,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.2 <- labeled_pca_plot(fit,3:4,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.3 <- labeled_pca_plot(fit,5:6,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.4 <- labeled_pca_plot(fit,7:8,samples$label,font_size = 7,
                       legend_label = "Cell labels")

plot_grid(p.pca3.1,p.pca3.2,p.pca3.3,p.pca3.4)
```

#### k = 12
Load the $k = 12$ Poisson NMF fit.
```{r load-unbinarized-fit-k12}
fit <- readRDS(file.path(out.dir, "/fit-Buenrostro2018-scd-ex-k=12.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-unbinarized-k12-1, fig.width=7, fig.height=4.5}
p.pca1.1 <- pca_plot(poisson2multinom(fit),pcs = 1:2,fill = "none")
p.pca1.2 <- pca_plot(poisson2multinom(fit),pcs = 3:4,fill = "none")
p.pca1.3 <- pca_plot(poisson2multinom(fit),pcs = 5:6,fill = "none")
p.pca1.4 <- pca_plot(poisson2multinom(fit),pcs = 7:8,fill = "none")

plot_grid(p.pca1.1,p.pca1.2,p.pca1.3,p.pca1.4)
```

Some of the structure is more evident from “hexbin” plots 
showing the density of the points.

```{r pca-unbinarized-k12-2, fig.width=7, fig.height=4.5}
breaks <- c(0,1,5,10,100,Inf)
p.pca2.1 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 1:2, breaks = breaks) + guides(fill = "none")
p.pca2.2 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 3:4, breaks = breaks) + guides(fill = "none")
p.pca2.3 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 5:6, breaks = breaks) + guides(fill = "none")
p.pca2.4 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 7:8, breaks = breaks) + guides(fill = "none")

plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4)
```

Next, we layer the cell labels onto the PC plots.

```{r pca-unbinarized-k12-3, fig.width=8, fig.height=6}
p.pca3.1 <- labeled_pca_plot(fit,1:2,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.2 <- labeled_pca_plot(fit,3:4,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.3 <- labeled_pca_plot(fit,5:6,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.4 <- labeled_pca_plot(fit,7:8,samples$label,font_size = 7,
                       legend_label = "Cell labels")

plot_grid(p.pca3.1,p.pca3.2,p.pca3.3,p.pca3.4)
```

#### k = 13
Load the $k = 13$ Poisson NMF fit.
```{r load-unbinarized-fit-k13}
fit <- readRDS(file.path(out.dir, "/fit-Buenrostro2018-scd-ex-k=13.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-unbinarized-k13-1, fig.width=7, fig.height=4.5}
p.pca1.1 <- pca_plot(poisson2multinom(fit),pcs = 1:2,fill = "none")
p.pca1.2 <- pca_plot(poisson2multinom(fit),pcs = 3:4,fill = "none")
p.pca1.3 <- pca_plot(poisson2multinom(fit),pcs = 5:6,fill = "none")
p.pca1.4 <- pca_plot(poisson2multinom(fit),pcs = 7:8,fill = "none")

plot_grid(p.pca1.1,p.pca1.2,p.pca1.3,p.pca1.4)
```

Some of the structure is more evident from “hexbin” plots 
showing the density of the points.

```{r pca-unbinarized-k13-2, fig.width=7, fig.height=4.5}
breaks <- c(0,1,5,10,100,Inf)
p.pca2.1 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 1:2, breaks = breaks) + guides(fill = "none")
p.pca2.2 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 3:4, breaks = breaks) + guides(fill = "none")
p.pca2.3 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 5:6, breaks = breaks) + guides(fill = "none")
p.pca2.4 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 7:8, breaks = breaks) + guides(fill = "none")

plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4)
```

Next, we layer the cell labels onto the PC plots.

```{r pca-unbinarized-k13-3, fig.width=8, fig.height=6}
p.pca3.1 <- labeled_pca_plot(fit,1:2,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.2 <- labeled_pca_plot(fit,3:4,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.3 <- labeled_pca_plot(fit,5:6,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.4 <- labeled_pca_plot(fit,7:8,samples$label,font_size = 7,
                       legend_label = "Cell labels")

plot_grid(p.pca3.1,p.pca3.2,p.pca3.3,p.pca3.4)
```


#### k = 14
Load the $k = 14$ Poisson NMF fit.
```{r load-unbinarized-fit-k14}
fit <- readRDS(file.path(out.dir, "/fit-Buenrostro2018-scd-ex-k=14.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-unbinarized-k14-1, fig.width=7, fig.height=4.5}
p.pca1.1 <- pca_plot(poisson2multinom(fit),pcs = 1:2,fill = "none")
p.pca1.2 <- pca_plot(poisson2multinom(fit),pcs = 3:4,fill = "none")
p.pca1.3 <- pca_plot(poisson2multinom(fit),pcs = 5:6,fill = "none")
p.pca1.4 <- pca_plot(poisson2multinom(fit),pcs = 7:8,fill = "none")

plot_grid(p.pca1.1,p.pca1.2,p.pca1.3,p.pca1.4)
```

Some of the structure is more evident from “hexbin” plots 
showing the density of the points.

```{r pca-unbinarized-k14-2, fig.width=7, fig.height=4.5}
breaks <- c(0,1,5,10,100,Inf)
p.pca2.1 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 1:2, breaks = breaks) + guides(fill = "none")
p.pca2.2 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 3:4, breaks = breaks) + guides(fill = "none")
p.pca2.3 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 5:6, breaks = breaks) + guides(fill = "none")
p.pca2.4 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 7:8, breaks = breaks) + guides(fill = "none")

plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4)
```

Next, we layer the cell labels onto the PC plots.

```{r pca-unbinarized-k14-3, fig.width=8, fig.height=6}
p.pca3.1 <- labeled_pca_plot(fit,1:2,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.2 <- labeled_pca_plot(fit,3:4,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.3 <- labeled_pca_plot(fit,5:6,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.4 <- labeled_pca_plot(fit,7:8,samples$label,font_size = 7,
                       legend_label = "Cell labels")

plot_grid(p.pca3.1,p.pca3.2,p.pca3.3,p.pca3.4)
```


#### k = 15
Load the $k = 15$ Poisson NMF fit.
```{r load-unbinarized-fit-k15}
fit <- readRDS(file.path(out.dir, "/fit-Buenrostro2018-scd-ex-k=15.rds"))$fit
```

Plot PCs of the topic proportions.

```{r pca-unbinarized-k15-1, fig.width=7, fig.height=4.5}
p.pca1.1 <- pca_plot(poisson2multinom(fit),pcs = 1:2,fill = "none")
p.pca1.2 <- pca_plot(poisson2multinom(fit),pcs = 3:4,fill = "none")
p.pca1.3 <- pca_plot(poisson2multinom(fit),pcs = 5:6,fill = "none")
p.pca1.4 <- pca_plot(poisson2multinom(fit),pcs = 7:8,fill = "none")

plot_grid(p.pca1.1,p.pca1.2,p.pca1.3,p.pca1.4)
```

Some of the structure is more evident from “hexbin” plots 
showing the density of the points.

```{r pca-unbinarized-k15-2, fig.width=7, fig.height=4.5}
breaks <- c(0,1,5,10,100,Inf)
p.pca2.1 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 1:2, breaks = breaks) + guides(fill = "none")
p.pca2.2 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 3:4, breaks = breaks) + guides(fill = "none")
p.pca2.3 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 5:6, breaks = breaks) + guides(fill = "none")
p.pca2.4 <- pca_hexbin_plot(poisson2multinom(fit), pcs = 7:8, breaks = breaks) + guides(fill = "none")

plot_grid(p.pca2.1,p.pca2.2,p.pca2.3,p.pca2.4)
```

Next, we layer the cell labels onto the PC plots.

```{r pca-unbinarized-k15-3, fig.width=8, fig.height=6}
p.pca3.1 <- labeled_pca_plot(fit,1:2,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.2 <- labeled_pca_plot(fit,3:4,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.3 <- labeled_pca_plot(fit,5:6,samples$label,font_size = 7,
                       legend_label = "Cell labels")
p.pca3.4 <- labeled_pca_plot(fit,7:8,samples$label,font_size = 7,
                       legend_label = "Cell labels")

plot_grid(p.pca3.1,p.pca3.2,p.pca3.3,p.pca3.4)
```
