---
title: "Exploring Cicero co-assessibility data in kidney data"
author: Peter Carbonetto
output: workflowr::wflow_html
---

The aim of this short analysis is to get a better understanding of the
Cicero co-accessibility data, and how (and whether) these data can be
used to connect chromatin accessibility peaks to genes in order to
identify "driving genes" for topics. As an illustration, here I focus
on the Cicero data for a single gene, *Slc12a1*, that was highlighted
in the [Cusanovich et al (2018) *Cell* paper][cusanovich-2018-cell] in
connection with the "loop of henle" cell type (see Fig. 5 of that
paper).

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages used in the analysis below.

```{r load-pkgs, message=FALSE}
library(fastTopics)
library(ggplot2)
library(cowplot)
library(ashr)
```

Load the base-pair positions of the genes for the mm9 Mouse Genome
Assembly.

```{r load-genes}
load("data/mm9_seq_gene.RData")
```

Load the cicero co-accessibility data for gene *Slc12a1.*

TO DO: Explain where these data were downloaded from and how they were
compiled.

```{r load-cicero}
load("data/Cusanovich_2018/processed_data/slc12a1_data.RData")
cicero <- transform(cicero,
                    Peak1 = as.character(Peak1),
                    Peak2 = as.character(Peak2))
```

[cusanovich-2018-cell]: https://doi.org/10.1016/j.cell.2018.06.052
