---
title: "Gene analysis using topic modeling results for Cusanovich *et al* (2018) scATAC-seq data"
author: Kaixuan Luo
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: false
---

Here we perform gene analysis
for the Cusanovich *et al* (2018) scATAC-seq result inferred 
from the multinomial topic model with $k = 13$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = FALSE,results = "hold",
                      fig.align = "center",dpi = 120, message = FALSE, warning = FALSE)
```

## Load packages and some functions used in this analysis
```{r load-pkgs}
library(Matrix)
library(fastTopics)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(plotly)
library(htmlwidgets)
library(DT)
library(reshape)
source("code/plots.R")
```

## Load samples and topic model ($k = 13$) results.
```{r load-samples}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018/"
samples <- readRDS(paste0(out.dir, "/samples-clustering-Cusanovich2018.rds"))
```

```{r load-fit}
fit.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018/"
fit <- readRDS(file.path(fit.dir, "/fit-Cusanovich2018-scd-ex-k=13.rds"))$fit
fit_multinom <- poisson2multinom(fit)
```

## Visualize by Structure plot grouped by tissues
```{r structure-plot-tissue, fig.width=8, fig.height=2.5, message=FALSE}
set.seed(10)
colors_topics <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
                   "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928",
                   "gray")
rows <- sample(nrow(fit$L),4000)
samples$tissue <- as.factor(samples$tissue)

p.structure <- structure_plot(select(fit_multinom,loadings = rows),
                              grouping = samples[rows, "tissue"],n = Inf,gap = 40,
                              perplexity = 50,topics = 1:13,colors = colors_topics,
                              num_threads = 4,verbose = FALSE)

print(p.structure)
```

Load the clustering results
```{r structure-plot-kmeans,  fig.width=8, fig.height=2.5, message=FALSE}
set.seed(10)
rows <- sample(nrow(fit$L),4000)

p.structure.kmeans <- structure_plot(select(fit_multinom,loadings = rows),
                                     grouping = samples$cluster_kmeans[rows],n = Inf,gap = 40,
                                     perplexity = 50,topics = 1:13,colors = colors_topics,
                                     num_threads = 4,verbose = FALSE)
print(p.structure.kmeans)
```

Distribution of tissue labels by cluster.
```{r freq-table-cluster-tissue, fig.width=6, fig.height=4}
freq_table_cluster_tissue <- with(samples,table(tissue,cluster_kmeans))

freq_table_cluster_tissue <- as.data.frame.matrix(freq_table_cluster_tissue)
# DT::datatable(freq_table_cluster_tissue, 
#               options = list(pageLength = nrow(freq_table_cluster_tissue)), 
#               rownames = T, caption = "Number of cells")

create_celllabel_cluster_heatmap(samples$tissue, samples$cluster_kmeans, normalize_by = "column")

```

Distribution of cell labels by cluster.
```{r freq-table-cluster-celllabel, fig.width=6, fig.height=10}
freq_table_cluster_celllabel <- with(samples,table(cell_label,cluster_kmeans))

freq_table_cluster_celllabel <- as.data.frame.matrix(freq_table_cluster_celllabel)
# DT::datatable(freq_table_cluster_celllabel, 
#               options = list(pageLength = nrow(freq_table_cluster_celllabel)), 
#               rownames = T, caption = "Number of cells")

create_celllabel_cluster_heatmap(samples$cell_label, samples$cluster_kmeans, normalize_by = "column")
```

Top 5 cell types 
```{r}
top_celltypes_table <- data.frame(matrix(nrow=5, ncol = ncol(freq_table_cluster_celllabel)))
colnames(top_celltypes_table) <- colnames(freq_table_cluster_celllabel)
for (k in 1:ncol(freq_table_cluster_celllabel)){
  top_celltypes <- rownames(freq_table_cluster_celllabel)[head(order(freq_table_cluster_celllabel[,k], decreasing=TRUE), 5)]
  freq_top_celltypes <- freq_table_cluster_celllabel[top_celltypes, k]
  percent_top_celltypes <- freq_table_cluster_celllabel[top_celltypes, k]/sum(freq_table_cluster_celllabel[,k])
  top_celltypes_table[,k] <- sprintf("%s (%.1f%%)", top_celltypes, percent_top_celltypes*100)
}

DT::datatable(top_celltypes_table, rownames = T, caption = "Top 5 cell types in each cluster")

```

We can see the major cell types in the clusters (topics):

 * cluster 1  (topic 1):  Erythroblasts 
 * cluster 2  (topic 3):  Endothelial cells
 * cluster 3  (topic 7):  Cardiomyocytes cells
 * cluster 4  (topic 6):  B cells, Monocytes, Dendritic cells
 * cluster 5  (topic 9):  Astrocytes, Oligodendrocytes 
 * cluster 8  (topic 5):  Proximal tubule
 * cluster 10 (topic 4):  a mixture of pneumocytes, Loop of henle, Enterocytes, DCT/CD
 * cluster 11 (topic 12): T cells
 * cluster 13 (topic 2):  Cerebellar granule cells 
 * cluster 14 (topic 11): Ex. neurons and Inhibitory neurons
 * cluster 15 (topic 8):  Hepatocytes
 
## Gene score analysis

```{r set-dir}
# Set output directorry

fig.dir <- "output/plotly/Cusanovich2018"
dir.create(fig.dir, showWarnings = F, recursive = T)
```

### TSS model

Gene scores were computed using TSS based method as in Lareau *et al* Nature Biotech, 2019 as well as the `model 21` in `archR` paper.
This model weights chromatin accessibility around gene promoters by using bi-directional exponential decays from the TSS.

* Top genes
```{r top-genes-tss}
gene.dir <- paste0(out.dir, "/geneanalysis-Cusanovich2018-k=13-TSS-sum")
cat(sprintf("Directory of gene analysis result: %s \n", gene.dir))
genescore_res_tss <- readRDS(file.path(gene.dir, "genescore_result_topics.rds"))
genescore_res <- genescore_res_tss

genes <- genescore_res$genes
gene_mean_acc <- genescore_res$colmeans
gene_scores <- genescore_res$Z
gene_logFC <- genescore_res$beta

topics <- colnames(gene_scores)
top_genes <- data.frame(matrix(nrow=10, ncol = ncol(gene_scores)))
colnames(top_genes) <- topics

for (k in topics){
  top_genes[,k] <- genes$SYMBOL[head(order(abs(gene_scores[,k]), decreasing=TRUE), 10)]
}

DT::datatable(data.frame(rank = 1:10, top_genes), rownames = F, caption = "Top 10 genes by abs(gene z-scores)")

```

* Volcano plots
```{r gene-volcano-plots-tss, eval=FALSE, include=FALSE}
genescore_volcano_plot(genescore_res, label_above_quantile = 0.99,
                       labels = genescore_res$genes$SYMBOL, max.overlaps = 20,
                       subsample_below_quantile = 0.5, subsample_rate = 0.1)
```

Explore the volcano plots [interactively][volcano-plotly-tss-k1]
```{r gene-volcano-plotly-tss-1}
for ( k in 1:ncol(genescore_res$beta) ){
  p.volcano.plotly <- genescore_volcano_plotly(genescore_res,k,
                                             file = sprintf("%s/volcano_topic_%s_%s.html", fig.dir, k, "tss-sum"),
                                             labels = genescore_res$genes$SYMBOL)
}
```

**Topic 1** (Erythroblasts)
```{r gene-volcano-plot-tss-1, fig.width=5, fig.height=4}
k <- 1
genescore_volcano_plot(genescore_res, k, label_above_quantile = 0.99,
                       labels = genescore_res$genes$SYMBOL, max.overlaps = 20,
                       subsample_below_quantile = 0.5, subsample_rate = 0.1)

```

Check some known marker genes
```{r marker-gene-scores-tss-1, fig.width=8, fig.height=4}
marker_genes <- c("Hbb-b1", "Hbb-b2", "Gypa")

gene_scores <- genescore_res$Z
rownames(gene_scores) <- genescore_res$genes$SYMBOL
marker_gene_scores <- gene_scores[grep(paste(sprintf("^%s$", marker_genes), collapse = "|"), rownames(gene_scores), ignore.case = T),]

par(mfrow = c(ceiling(nrow(marker_gene_scores)/2),2))
for(i in 1:nrow(marker_gene_scores)){
  barplot(marker_gene_scores[i,], xlab = "topics", ylab = "gene score", main = rownames(marker_gene_scores)[i], col = colors_topics)
}

```

**Topic 3** (Endothelial cells)
```{r gene-volcano-plot-tss-3, fig.width=5, fig.height=4}
k <- 3
genescore_volcano_plot(genescore_res, k, label_above_quantile = 0.99,
                       labels = genescore_res$genes$SYMBOL, max.overlaps = 20,
                       subsample_below_quantile = 0.5, subsample_rate = 0.1)

```

Check some known marker genes
```{r marker-gene-scores-tss-3, fig.width=8, fig.height=4}
marker_genes <- c("PECAM1", "CD106", "CD62E", "Sele", "Kdr", "ENG")
gene_scores <- genescore_res$Z
rownames(gene_scores) <- genescore_res$genes$SYMBOL
marker_gene_scores <- gene_scores[grep(paste(sprintf("^%s$", marker_genes), collapse = "|"), rownames(gene_scores), ignore.case = T),]

par(mfrow = c(ceiling(nrow(marker_gene_scores)/2),2))
for(i in 1:nrow(marker_gene_scores)){
  barplot(marker_gene_scores[i,], xlab = "topics", ylab = "gene score", main = rownames(marker_gene_scores)[i], col = colors_topics)
}

```

**Topic 5** (Proximal tubule)
```{r gene-volcano-plot-tss-5, fig.width=5, fig.height=4}
k <- 5
genescore_volcano_plot(genescore_res, k, label_above_quantile = 0.99,
                       labels = genescore_res$genes$SYMBOL, max.overlaps = 20,
                       subsample_below_quantile = 0.5, subsample_rate = 0.1)

```

Check some known marker genes
```{r marker-gene-scores-tss-5, fig.width=8, fig.height=4}
marker_genes <- c("PALDOB", "CUBN", "LRP2", "SLC34A1")
gene_scores <- genescore_res$Z
rownames(gene_scores) <- genescore_res$genes$SYMBOL
marker_gene_scores <- gene_scores[grep(paste(sprintf("^%s$", marker_genes), collapse = "|"), rownames(gene_scores), ignore.case = T),]

par(mfrow = c(ceiling(nrow(marker_gene_scores)/2),2))
for(i in 1:nrow(marker_gene_scores)){
  barplot(marker_gene_scores[i,], xlab = "topics", ylab = "gene score", main = rownames(marker_gene_scores)[i], col = colors_topics)
}

```

**Topic 7** (Cardiomyocytes)
```{r gene-volcano-plot-tss-7, fig.width=5, fig.height=4}
k <- 7
genescore_volcano_plot(genescore_res, k, label_above_quantile = 0.99,
                       labels = genescore_res$genes$SYMBOL, max.overlaps = 20,
                       subsample_below_quantile = 0.5, subsample_rate = 0.1)

```

Check some known marker genes
```{r marker-gene-scores-tss-7, fig.width=8, fig.height=8}
marker_genes <- c("Nppa", "Myl4", "SLN", "PITX2", "Myl7", "Gja5", "Myl2", "Myl3", "IRX4", "HAND1", "HEY2")
gene_scores <- genescore_res$Z
rownames(gene_scores) <- genescore_res$genes$SYMBOL
marker_gene_scores <- gene_scores[grep(paste(sprintf("^%s$", marker_genes), collapse = "|"), rownames(gene_scores), ignore.case = T),]

par(mfrow = c(ceiling(nrow(marker_gene_scores)/3),3))
for(i in 1:nrow(marker_gene_scores)){
  barplot(marker_gene_scores[i,], xlab = "topics", ylab = "gene score", main = rownames(marker_gene_scores)[i], col = colors_topics)
}

```

**Topic 8** (Hepatocytes)
```{r gene-volcano-plot-tss-8, fig.width=5, fig.height=4}
k <- 8
genescore_volcano_plot(genescore_res, k, label_above_quantile = 0.99,
                       labels = genescore_res$genes$SYMBOL, max.overlaps = 20,
                       subsample_below_quantile = 0.5, subsample_rate = 0.1)

```

Check some known marker genes
```{r marker-gene-scores-8, fig.width=8, fig.height=8}
marker_genes <- c("SERPINA1", "TTR", "ALB","AFP","CYP3A4","CYP7A1","FABP1","ALR","Glut1","MET","FoxA1","FoxA2","CD29","PTP4A2","Prox1", "HNF1B")
gene_scores <- genescore_res$Z
rownames(gene_scores) <- genescore_res$genes$SYMBOL
marker_gene_scores <- gene_scores[grep(paste(sprintf("^%s$", marker_genes), collapse = "|"), rownames(gene_scores), ignore.case = T),]

par(mfrow = c(ceiling(nrow(marker_gene_scores)/3),3))
for(i in 1:nrow(marker_gene_scores)){
  barplot(marker_gene_scores[i,], xlab = "topics", ylab = "gene score", main = rownames(marker_gene_scores)[i], col = colors_topics)
}

```

### Gene body model

Gene scores were computed using the gene score model (model 42) in the `archR` paper with some modifications.
This model uses bi-directional exponential decays from the gene TSS (extended upstream by 5 kb by default)
and the gene transcription termination site (TTS).
Note: the current version of the function does not account for neighboring gene boundaries.

* Top genes
```{r top-genes-genebody}
gene.dir <- paste0(out.dir, "/geneanalysis-Cusanovich2018-k=13-genebody-sum")
cat(sprintf("Directory of gene analysis result: %s \n", gene.dir))
genescore_res_gb <- readRDS(file.path(gene.dir, "genescore_result_topics.rds"))
genescore_res <- genescore_res_gb

genes <- genescore_res$genes
gene_mean_acc <- genescore_res$colmeans
gene_scores <- genescore_res$Z
gene_logFC <- genescore_res$beta

topics <- colnames(gene_scores)
top_genes <- data.frame(matrix(nrow=10, ncol = ncol(gene_scores)))
colnames(top_genes) <- topics

for (k in topics){
  top_genes[,k] <- genes$SYMBOL[head(order(abs(gene_scores[,k]), decreasing=TRUE), 10)]
}

DT::datatable(data.frame(rank = 1:10, top_genes), rownames = F, caption = "Top 10 genes by abs(gene z-scores)")

```

* Volcano plots
```{r gene-volcano-plots-genebody, eval=FALSE, include=FALSE}
genescore_volcano_plot(genescore_res, label_above_quantile = 0.99,
                       labels = genescore_res$genes$SYMBOL, max.overlaps = 20,
                       subsample_below_quantile = 0.5, subsample_rate = 0.1)
```

Explore the volcano plot [interactively][volcano-plotly-genebody-k1]
```{r gene-volcano-plotly-genebody-1}
for ( k in 1:ncol(genescore_res$beta) ){
  p.volcano.plotly <- genescore_volcano_plotly(genescore_res,k,
                                             file = sprintf("%s/volcano_topic_%s_%s.html", fig.dir, k, "genebody-sum"),
                                             labels = genescore_res$genes$SYMBOL)
}
```


**Topic 1** (Erythroblasts)
```{r gene-volcano-plot-genebody-1, fig.width=5, fig.height=4}
k <- 1
genescore_volcano_plot(genescore_res, k, label_above_quantile = 0.99,
                       labels = genescore_res$genes$SYMBOL, max.overlaps = 20,
                       subsample_below_quantile = 0.5, subsample_rate = 0.1)

```

Check some known marker genes
```{r marker-gene-scores-genebody-1, fig.width=8, fig.height=4}
marker_genes <- c("Hbb-b1", "Hbb-b2", "Gypa")

gene_scores <- genescore_res$Z
rownames(gene_scores) <- genescore_res$genes$SYMBOL
marker_gene_scores <- gene_scores[grep(paste(sprintf("^%s$", marker_genes), collapse = "|"), rownames(gene_scores), ignore.case = T),]

par(mfrow = c(ceiling(nrow(marker_gene_scores)/2),2))
for(i in 1:nrow(marker_gene_scores)){
  barplot(marker_gene_scores[i,], xlab = "topics", ylab = "gene score", main = rownames(marker_gene_scores)[i], col = colors_topics)
}

```

**Topic 3** (Endothelial cells)
```{r gene-volcano-plot-genebody-3, fig.width=5, fig.height=4}
k <- 3
genescore_volcano_plot(genescore_res, k, label_above_quantile = 0.99,
                       labels = genescore_res$genes$SYMBOL, max.overlaps = 20,
                       subsample_below_quantile = 0.5, subsample_rate = 0.1)

```

Check some known marker genes
```{r marker-gene-scores-genebody-3, fig.width=8, fig.height=4}
marker_genes <- c("PECAM1", "CD106", "CD62E", "Sele", "Kdr", "ENG")
gene_scores <- genescore_res$Z
rownames(gene_scores) <- genescore_res$genes$SYMBOL
marker_gene_scores <- gene_scores[grep(paste(sprintf("^%s$", marker_genes), collapse = "|"), rownames(gene_scores), ignore.case = T),]

par(mfrow = c(ceiling(nrow(marker_gene_scores)/2),2))
for(i in 1:nrow(marker_gene_scores)){
  barplot(marker_gene_scores[i,], xlab = "topics", ylab = "gene score", main = rownames(marker_gene_scores)[i], col = colors_topics)
}

```

**Topic 5** (Proximal tubule)
```{r gene-volcano-plot-genebody-5, fig.width=5, fig.height=4}
k <- 5
genescore_volcano_plot(genescore_res, k, label_above_quantile = 0.99,
                       labels = genescore_res$genes$SYMBOL, max.overlaps = 20,
                       subsample_below_quantile = 0.5, subsample_rate = 0.1)

```

Check some known marker genes
```{r marker-gene-scores-genebody-5, fig.width=8, fig.height=4}
marker_genes <- c("PALDOB", "CUBN", "LRP2", "SLC34A1")
gene_scores <- genescore_res$Z
rownames(gene_scores) <- genescore_res$genes$SYMBOL
marker_gene_scores <- gene_scores[grep(paste(sprintf("^%s$", marker_genes), collapse = "|"), rownames(gene_scores), ignore.case = T),]

par(mfrow = c(ceiling(nrow(marker_gene_scores)/2),2))
for(i in 1:nrow(marker_gene_scores)){
  barplot(marker_gene_scores[i,], xlab = "topics", ylab = "gene score", main = rownames(marker_gene_scores)[i], col = colors_topics)
}

```

**Topic 7** (Cardiomyocytes)
```{r gene-volcano-plot-genebody-7, fig.width=5, fig.height=4}
k <- 7
genescore_volcano_plot(genescore_res, k, label_above_quantile = 0.99,
                       labels = genescore_res$genes$SYMBOL, max.overlaps = 20,
                       subsample_below_quantile = 0.5, subsample_rate = 0.1)

```

Check some known marker genes
```{r marker-gene-scores-genebody-7, fig.width=8, fig.height=8}
marker_genes <- c("Nppa", "Myl4", "SLN", "PITX2", "Myl7", "Gja5", "Myl2", "Myl3", "IRX4", "HAND1", "HEY2")
gene_scores <- genescore_res$Z
rownames(gene_scores) <- genescore_res$genes$SYMBOL
marker_gene_scores <- gene_scores[grep(paste(sprintf("^%s$", marker_genes), collapse = "|"), rownames(gene_scores), ignore.case = T),]

par(mfrow = c(ceiling(nrow(marker_gene_scores)/3),3))
for(i in 1:nrow(marker_gene_scores)){
  barplot(marker_gene_scores[i,], xlab = "topics", ylab = "gene score", main = rownames(marker_gene_scores)[i], col = colors_topics)
}

```

**Topic 8** (Hepatocytes)
```{r gene-volcano-plot-genebody-8, fig.width=5, fig.height=4}
k <- 8
genescore_volcano_plot(genescore_res, k, label_above_quantile = 0.99,
                       labels = genescore_res$genes$SYMBOL, max.overlaps = 20,
                       subsample_below_quantile = 0.5, subsample_rate = 0.1)

```

Check some known marker genes for Hepatocytes
```{r marker-gene-scores-genebody-8, fig.width=8, fig.height=8}
marker_genes <- c("SERPINA1", "TTR", "ALB","AFP","CYP3A4","CYP7A1","FABP1","ALR","Glut1","MET","FoxA1","FoxA2","CD29","PTP4A2","Prox1", "HNF1B")
gene_scores <- genescore_res$Z
rownames(gene_scores) <- genescore_res$genes$SYMBOL
marker_gene_scores <- gene_scores[grep(paste(sprintf("^%s$", marker_genes), collapse = "|"), rownames(gene_scores), ignore.case = T),]

par(mfrow = c(ceiling(nrow(marker_gene_scores)/3),3))
for(i in 1:nrow(marker_gene_scores)){
  barplot(marker_gene_scores[i,], xlab = "topics", ylab = "gene score", main = rownames(marker_gene_scores)[i], col = colors_topics)
}

```

### Compare gene scores from the gene-body model vs. the TSS model.
```{r compare-genebody-tss, fig.width=8, fig.height=7}
m     <- ncol(genescore_res_gb$Z)
plots <- vector("list",m)
names(plots) <- colnames(genescore_res_gb$Z)
for (i in 1:m) {
  dat <- data.frame(genebody = genescore_res_gb$Z[,i], tss = genescore_res_tss$Z[,i])
  plots[[i]] <- 
    ggplot(dat,aes_string(x = "genebody",y = "tss")) +
    geom_point(shape = 21, na.rm = TRUE, size = 1, alpha = 1/10) +
    geom_abline(intercept = 0, slope = 1, color="blue") + 
    labs(x = "gene body model",y = "TSS model", 
         title = paste("topic",i)) +
    theme_cowplot(9)
}

do.call(plot_grid,plots)

```

## Gene-set enrichment analysis (GSEA)

### TSS model
Top gene sets/pathways
```{r top-genesets-tss}
gene.dir <- paste0(out.dir, "/geneanalysis-Cusanovich2018-k=13-TSS-sum")
cat(sprintf("Directory of gene analysis result: %s \n", gene.dir))
load(file.path(gene.dir, "genescores_gsea.Rdata"))

top_pathways_up <- top_pathways_down <- data.frame(matrix(nrow=10, ncol = ncol(gsea_res$pval)))
colnames(top_pathways_up) <- colnames(top_pathways_down) <- colnames(gsea_res$pval)

for (k in 1:ncol(gsea_res$pval)){
  gsea_topic <- data.frame(pathway = rownames(gsea_res$pval),  
                           pval = gsea_res$pval[,k],
                           log2err = gsea_res$log2err[,k],
                           ES = gsea_res$ES[,k],
                           NES = gsea_res$NES[,k])
  gsea_up <- gsea_topic[gsea_topic$ES > 0,]
  top_IDs_up <- as.character(gsea_up[head(order(gsea_up$pval), 10), "pathway"])
  top_IDs_up <- gene_set_info[match(top_IDs_up, gene_set_info$id),c("name", "id")]
  top_pathways_up[,k] <- paste0(top_IDs_up$name, "(", top_IDs_up$id, ")")
  
  gsea_down <- gsea_topic[gsea_topic$ES < 0,]
  top_IDs_down <- as.character(gsea_down[head(order(gsea_down$pval), 10), "pathway"])
  top_IDs_down <- gene_set_info[match(top_IDs_down, gene_set_info$id),c("name", "id")]
  top_pathways_down[,k] <- paste0(top_IDs_down$name, "(", top_IDs_down$id, ")")
  
}

DT::datatable(data.frame(rank = 1:10, top_pathways_up), rownames = F,
              caption = "Top 10 pathways enriched at the top of the gene rank list.")

```

* Scatterplot of GSEA results
**Topic 1** (highlighting "Erythroblast" cells)
```{r gsea-plot-tss-1, fig.width=8, fig.height=7}
k   <- 1
name_interest <- "erythroblast"
label_geneset_ids <- gene_set_info[gsea_res$pval[,k] < 1e-6 & grepl(name_interest, gene_set_info$name, ignore.case = T), "id"]
cat(sprintf("highlighting %d %s gene sets \n", length(label_geneset_ids), name_interest))

create_genescore_gsea_plot(gene_set_info, gsea_res, k, label_geneset_ids, 
                                     title = sprintf("topic %d GSEA plot",k))

if(length(label_geneset_ids) > 0)
   DT::datatable(gene_set_info[gene_set_info$id %in% label_geneset_ids, 
                               c("name", "id", "database", "category_code", "description_brief")],  
                 rownames = F)

```

Explore the GSEA plot [interactively][gsea-plotly-tss-k1]
```{r gsea-plotly-tss-1}
for ( k in 1:ncol(gsea_res$pval)){
  p.gsea.plotly <- create_genescore_gsea_plotly(gene_set_info, gsea_res, k)
  saveWidget(p.gsea.plotly,file = sprintf("%s/gsea_topic_%s_%s.html", fig.dir, k, "tss-sum"),
             selfcontained = TRUE,title = "GSEA")
}
```

### Gene body model
* Top gene sets
```{r top-genesets-genebody}
gene.dir <- paste0(out.dir, "/geneanalysis-Cusanovich2018-k=13-genebody-sum")
cat(sprintf("Directory of gene analysis result: %s \n", gene.dir))
load(file.path(gene.dir, "genescores_gsea.Rdata"))

top_pathways_up <- top_pathways_down <- data.frame(matrix(nrow=10, ncol = ncol(gsea_res$pval)))
colnames(top_pathways_up) <- colnames(top_pathways_down) <- colnames(gsea_res$pval)

for (k in 1:ncol(gsea_res$pval)){
  gsea_topic <- data.frame(pathway = rownames(gsea_res$pval),  
                           pval = gsea_res$pval[,k],
                           log2err = gsea_res$log2err[,k],
                           ES = gsea_res$ES[,k],
                           NES = gsea_res$NES[,k])
  gsea_up <- gsea_topic[gsea_topic$ES > 0,]
  top_IDs_up <- as.character(gsea_up[head(order(gsea_up$pval), 10), "pathway"])
  top_pathways_up[,k] <- gene_set_info[match(top_IDs_up, gene_set_info$id),c("name", "id")] %>% 
    unite("pathway", c("name", "id"), sep = ".", remove = TRUE)
  
  gsea_down <- gsea_topic[gsea_topic$ES < 0,]
  top_IDs_down <- as.character(gsea_down[head(order(gsea_down$pval), 10), "pathway"])
  top_pathways_down[,k] <- gene_set_info[match(top_IDs_down, gene_set_info$id),c("name", "id")] %>% 
    unite("pathway", c("name", "id"), sep = ".", remove = TRUE)
}

DT::datatable(data.frame(rank = 1:10, top_pathways_up), rownames = F,
              caption = "Top 10 pathways enriched at the top of the gene rank list.")

```

* Scatterplot of GSEA results
**Topic 1** (highlighting "Erythroblast" cells)
```{r gsea-plot-genebody-1, fig.width=8, fig.height=7}
k   <- 1
name_interest <- "erythroblast"
label_geneset_ids <- gene_set_info[gsea_res$pval[,k] < 1e-6 & grepl(name_interest, gene_set_info$name, ignore.case = T), "id"]
cat(sprintf("highlighting %d %s gene sets \n", length(label_geneset_ids), name_interest))

create_genescore_gsea_plot(gene_set_info, gsea_res, k, label_geneset_ids, 
                                     title = sprintf("topic %d GSEA plot",k))

if(length(label_geneset_ids) > 0)
   DT::datatable(gene_set_info[gene_set_info$id %in% label_geneset_ids, 
                               c("name", "id", "database", "category_code", "description_brief")],  
                 rownames = F)
```

Explore the GSEA plot [interactively][gsea-plotly-genebody-k1]
```{r gsea-plotly-genebody-1}
for ( k in 1:ncol(gsea_res$pval)){
  p.gsea.plotly <- create_genescore_gsea_plotly(gene_set_info, gsea_res, k)
saveWidget(p.gsea.plotly,file = sprintf("%s/gsea_topic_%s_%s.html", fig.dir, k, "genebody-sum"),
           selfcontained = TRUE,title = "GSEA")
}

```


[volcano-plotly-tss-k1]: plotly/Cusanovich2018/volcano_topic_1_tss-sum.html
[volcano-plotly-genebody-k1]: plotly/Cusanovich2018/volcano_topic_1_genebody-sum.html
[gsea-plotly-tss-k1]: plotly/Cusanovich2018/gsea_topic_1_tss-sum.html
[gsea-plotly-genebody-k1]: plotly/Cusanovich2018/gsea_topic_1_genebody-sum.html
