---
title: "Motif analysis using topic modeling results for Cusanovich *et al* (2018) scATAC-seq data"
author: Kaixuan Luo
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: false
---

Here we perform TF motif analysis
for the Cusanovich *et al* (2018) scATAC-seq result inferred 
from the multinomial topic model with $k = 13$.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = FALSE,results = "hold",
                      fig.align = "center",dpi = 120)
```

## Load packages and some functions used in this analysis
```{r load-pkgs, message=FALSE}
library(Matrix)
library(fastTopics)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(plotly)
library(htmlwidgets)
library(DT)
library(reshape2)
source("code/plots.R")
```


## Load data and topic model results
Load the data and the $k = 13$ Poisson NMF fit results.
```{r load-data}
data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Cusanovich_2018/processed_data/"
load(file.path(data.dir, "Cusanovich_2018.RData"))
rm(counts)
```

```{r load-fit}
fit.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018"
fit <- readRDS(file.path(fit.dir, "/fit-Cusanovich2018-scd-ex-k=13.rds"))$fit
fit_multinom <- poisson2multinom(fit)
```

## Visualize by Structure plot grouped by tissues
```{r structure-plot-tissue, fig.width=7, fig.height=2.5, message=FALSE}
set.seed(10)
colors_topics <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
                   "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928",
                   "gray")
rows <- sample(nrow(fit$L),4000)
samples$tissue <- as.factor(samples$tissue)

p.structure <- structure_plot(select(fit_multinom,loadings = rows),
                                grouping = samples[rows, "tissue"],n = Inf,gap = 40,
                                perplexity = 50,topics = 1:13,colors = colors_topics,
                                num_threads = 4,verbose = FALSE)

print(p.structure)
```

## Differential accessbility analysis of the ATAC-seq regions for the topics
Load results from differential accessbility analysis for the topics

```{r}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Cusanovich_2018"
cat(sprintf("Load results from %s \n", out.dir))
```

```{r load_diff_count_res}
diff_count_topics <- readRDS(file.path(out.dir, "/diffcount-Cusanovich2018-13topics.rds"))
```

Distribution of z-scores
```{r hist-zscore, fig.width=7, fig.height=5}

zscore_topics <-  melt(diff_count_topics$Z)
colnames(zscore_topics) <- c("region", "topic", "zscore")
levels(zscore_topics$topic) <- colnames(diff_count_topics$Z)

z.quantile.99 <- apply(abs(diff_count_topics$Z), 2, quantile, 0.99)
cat("z-score 99% quantile: \n")
print(z.quantile.99)

p.hist.zscores <- ggplot(zscore_topics, aes(x=zscore)) + 
  geom_histogram(binwidth=1, color="black", fill="white") + 
  coord_cartesian(xlim = c(-10, 30)) + theme_cowplot(font_size = 10) +
  facet_wrap(~ topic, ncol=4)

print(p.hist.zscores)
```

Volcano plot of the regions for topic 1
```{r region-volcano-plot}
volcano_plot(diff_count_topics,k = 1,label_above_quantile = Inf, 
             subsample_below_quantile = 0.8, subsample_rate = 0.1)
```

## Motif enrichment analysis using [HOMER](http://homer.ucsd.edu/homer/ngs/peakMotifs.html)

* Details about HOMER motif analysis:
    - http://homer.ucsd.edu/homer/motif/
    - http://homer.ucsd.edu/homer/ngs/peakMotifs.html

* Motif enrichment result using regions with z-score above 99% quantile.

Compile Homer results across topics
```{r compile-homer-motif-results}
homer.dir <- paste0(out.dir, "/motifanalysis-Cusanovich2018-k=13-quantile/HOMER/")
cat(sprintf("Directory of motif analysis result: %s \n", homer.dir))
homer_res_topics <- readRDS(file.path(homer.dir, "/homer_knownResults.rds"))
selected_regions <- readRDS(file.path(homer.dir, "/selected_regions.rds"))

# Compile Homer results (pvalue and ranking) across topics
motif_res <- compile_homer_motif_res(homer_res_topics)
saveRDS(motif_res, paste0(homer.dir, "/homer_motif_enrichment_results.rds"))

motif_table <- data.frame(motif = gsub("/.*", "", rownames(motif_res$mlog10P)),
                          motif_res$mlog10P)
DT::datatable(motif_table, rownames = F, caption = "Motif enrichment (-log10P)")
```

### Top 10 motifs
```{r top-motifs}
cat("Number of regions selected for each topic: \n")
print(mapply(nrow, selected_regions[1:(length(selected_regions)-1)]))

colnames_homer <- c("motif_name", "consensus", "P", "log10P", "Padj",  "num_target", "percent_target", "num_bg", "percent_bg")

top_motifs <- data.frame(matrix(nrow=10, ncol = length(homer_res_topics)))
colnames(top_motifs) <- names(homer_res_topics)
for (k in 1:length(homer_res_topics)){
  homer_res <- homer_res_topics[[k]]
  colnames(homer_res) <- colnames_homer
  homer_res <- homer_res %>% separate(motif_name, c("motif", "origin", "database"), "/")
  top_motifs[,k] <- head(homer_res$motif, 10)
}

DT::datatable(data.frame(rank = 1:10, top_motifs), rownames = F, caption = "Top 10 motifs enriched in each topic.")

```

### Heatmap of motif enrichment across topics

Clustering motifs by hierarchical clustering (motifs with similar enrichment across topics are plotted together)
```{r motif-heatmap-1, fig.height=10, fig.width=7}
create_motif_enrichment_heatmap(motif_res, cluster_motifs = TRUE, cluster_topics = FALSE, filter_motifs = TRUE, min_enrichment = 50, 
                                max_enrichment = 100, method_cluster = "average", font.size.motifs = 5, font.size.topics = 9)

```

Cluster both motifs and topics by hierarchical clustering
```{r motif-heatmap-2, fig.height=10, fig.width=7}
create_motif_enrichment_heatmap(motif_res, cluster_motifs = TRUE, cluster_topics = TRUE, filter_motifs = TRUE, min_enrichment = 50, 
                                max_enrichment = 100, method_cluster = "average", font.size.motifs = 5, font.size.topics = 9)

```

### Scatterplots of motif enrichment

Plot enrichment (-log10 p-value) and ranking of the motifs
```{r motif-enrichment-ranking-plot-1, warning=FALSE}
plots <- vector("list", ncol(motif_res$mlog10P))
names(plots) <- colnames(motif_res$mlog10P)

for( i in 1:length(plots)){
 plots[[i]] <- create_motif_enrichment_ranking_plot(motif_res, k = i, 
                                                    max.overlaps = 15, subsample = FALSE)
}

# do.call(plot_grid,plots)
```

Plot motif enrichment (-log10 p-value) and ranking in topic 1
```{r motif-enrichment-ranking-plot-2, warning=FALSE}
print(plots[[1]])
```

Plot motif enrichment (-log10 p-value) in each topic vs other topics
```{r motif-enrichment-plot-1, warning=FALSE}
plots <- vector("list", ncol(motif_res$mlog10P))
names(plots) <- colnames(motif_res$mlog10P)

for( i in 1:length(homer_res_topics)){
 plots[[i]] <- create_motif_enrichment_plot(motif_res, k = i, 
                                            max.overlaps = 15, subsample = TRUE)
}

# do.call(plot_grid,plots)

```

Plot motif enrichment (-log10 p-value) in topic 1 vs other topics
```{r motif-enrichment-plot-2, warning=FALSE}
print(plots[[1]])
```

### Motif enrichment vs gene score

Load pre-computed gene scores
```{r load-gene-scores}
gene.dir <- paste0(out.dir, "/geneanalysis-Cusanovich2018-k=13-TSS-sum")
cat(sprintf("Directory of gene analysis result: %s \n", gene.dir))
genescore_res <- readRDS(file.path(gene.dir, "genescore_result_topics.rds"))

genes <- genescore_res$genes
gene_scores <- genescore_res$Z
rownames(gene_scores) <- genes$SYMBOL
gene_logFC <- genescore_res$beta
rownames(gene_logFC) <- genes$SYMBOL

```

Get TF genes 
```{r get-TF-genes}
motif_names <- gsub("\\s*\\(.*", "", motif_res$motifs$motif)
gene_names <- genes$SYMBOL

TF_genes <- intersect(toupper(motif_names), toupper(gene_names))
cat(sprintf("%s TF genes mapped between motif names and gene symbol. \n", length(TF_genes)))

```

Compute correlation between motif enrichment and gene score:

Plot motif enrichment (-log10 p-value) and correlation to gene scores for topic 1
```{r motif-cor-plot}

motif_gene_mapping <- create_motif_enrichment_cor_plot(motif_res$mlog10P, gene_scores, motif_names, gene_names, TF_genes, k = 1, 
                                                       cor.method = "pearson", max.overlaps = 15)

motif_gene_mapping <- motif_gene_mapping[with(motif_gene_mapping, order(motif.mlog10P1*cor, decreasing = T)),]
head(motif_gene_mapping)
```

Plot motif enrichment (-log10 p-value) and gene scores for GATA genes
```{r motif-gene-plot-1}
GATA_genes <- grep("GATA", TF_genes, ignore.case=T, value=T)

plots <- create_motif_gene_scatterplot(motif_matrix = motif_res$mlog10P,
                                       gene_matrix = gene_scores, 
                                       motif_names, gene_names, 
                                       GATA_genes)

do.call(plot_grid,plots)
```

