---
title: "Motif analysis using topic modeling and DA results (v2) for Buenrostro *et al* (2018) scATAC-seq result"
author: Kaixuan Luo
output:
  workflowr::wflow_html:
    code_folding: hide
---

Here we perform TF motif analysis
for the Buenrostro *et al* (2018) scATAC-seq result inferred 
from the multinomial topic model with $k = 10$.

We use binarized data downloaded from original paper.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = FALSE,results = "hold",
                      fig.align = "center",dpi = 120)
```

## Load packages and some functions used in this analysis
```{r load-pkgs, message=FALSE}
library(Matrix)
library(fastTopics)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(plotly)
library(DT)
library(reshape2)
source("code/motif_analysis.R")
source("code/plots.R")
```

## Load data and topic model results
Data downloaded from original paper.
Load the **binarized** data and the $k = 10$ Poisson NMF fit results 
```{r load-data}
data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/processed_data/"
load(file.path(data.dir, "Buenrostro_2018_binarized_counts.RData"))
cat(sprintf("%d x %d counts matrix.\n",nrow(counts),ncol(counts)))
```

```{r load-fit}
fit.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Buenrostro_2018/binarized/"
fit <- readRDS(file.path(fit.dir, "/fit-Buenrostro2018-binarized-scd-ex-k=10.rds"))$fit
fit <- poisson2multinom(fit)
```

## Structure plot
```{r structure-plot, fig.width=7, fig.height=2.5, message=FALSE}
topic_colors <- c("darkorange","limegreen","magenta","gold","skyblue",
                  "darkblue","dodgerblue","darkmagenta","red","olivedrab")

set.seed(1)
# labels <- factor(samples$label, levels = c("HSC", "MPP", "CMP", "GMP", "mono", "MEP", "LMPP", "CLP", "pDC", "UNK"))

labels <- factor(samples$label, c("mono","pDC","MEP","HSC","MPP","CLP",
                                 "LMPP","CMP","GMP","UNK"))
structure_plot(fit,grouping = labels,colors = topic_colors,
               # topics = 1:10,
               gap = 20,perplexity = 50,verbose = FALSE)
```

## Motif enrichment analysis using HOMER

  * About HOMER motif analysis
    - http://homer.ucsd.edu/homer/motif/
    - http://homer.ucsd.edu/homer/ngs/peakMotifs.html

  * [Script for motif enrichment analysis using HOMER](../scripts/postfit_motif_analysis_Buenrostro2018.R)

### Motif enrichment result using regions with DA p-value < 0.05

Load and compile HOMER results across topics
```{r compile-homer-motif-results-p0.05}
postfit.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Buenrostro_2018/binarized/postfit_v2"
homer.dir <- paste0(postfit.dir, "/motifanalysis-Buenrostro2018-k=10/HOMER/DA_pval_0.05_regions")
cat(sprintf("Directory of motif analysis result: %s \n", homer.dir))
homer_res_topics <- readRDS(file.path(homer.dir, "/homer_knownResults.rds"))
selected_regions <- readRDS(file.path(homer.dir, "/selected_regions.rds"))

# Compile Homer results (pvalue and ranking) across topics
motif_res <- compile_homer_motif_res(homer_res_topics)
saveRDS(motif_res, paste0(homer.dir, "/homer_motif_enrichment_results.rds"))
cat("compiled homer motif results are saved in", paste0(homer.dir, "/homer_motif_enrichment_results.rds \n"))
```

```{r all-motif-enrichment-res-p0.05, eval=FALSE, include=FALSE}
motif_table <- data.frame(motif = gsub("/.*", "", rownames(motif_res$mlog10P)),
                          round(motif_res$mlog10P,2))
DT::datatable(motif_table, rownames = F, caption = "Motif enrichment (-log10P)")
```

Top 10 motifs for each topic
```{r top-motifs-p0.05}
cat("Number of regions selected for each topic: \n")
print(mapply(nrow, selected_regions[1:(length(selected_regions)-1)]))

colnames_homer <- c("motif_name", "consensus", "P", "log10P", "Padj",  "num_target", "percent_target", "num_bg", "percent_bg")

top_motifs <- data.frame(matrix(nrow=10, ncol = length(homer_res_topics)))
colnames(top_motifs) <- names(homer_res_topics)
for (k in 1:length(homer_res_topics)){
  homer_res <- homer_res_topics[[k]]
  colnames(homer_res) <- colnames_homer
  homer_res <- homer_res %>% separate(motif_name, c("motif", "origin", "database"), "/")
  top_motifs[,k] <- head(homer_res$motif, 10)
}

DT::datatable(data.frame(rank = 1:10, top_motifs), rownames = F, caption = "Top 10 motifs enriched in each topic.")

```

Heatmap of motif enrichment -log10(p-value). 
```{r motif-heatmap-pvalue-p0.05, fig.width=7, fig.height=12}
create_motif_enrichment_heatmap(motif_res, enrichment = "-log10(p-value)", 
                                cluster_motifs = TRUE, cluster_topics = TRUE, motif_filter = 10, horizontal = FALSE,
                                enrichment_range = c(0,100), method_cluster = "average", font.size.motifs = 4, font.size.topics = 9)
```

```{r selected-motifs-p0.05, include=FALSE, eval=FALSE, fig.width=7, fig.height=6}
# Heatmap of motif enrichment for selected TF motifs
toMatch <- c("^GATA\\d*$", "^CEBP.?$", "^SPI.?$", "^IRF\\d*$", "^STAT\\d*$", "^TCF\\d*$", "^BCL\\d*$", "^CTCF$", "^ERG$")
selected_motifs <- grep(paste(toMatch,collapse="|"), motif_res$motifs$motif, ignore.case = T, value = T)
rows <- match(selected_motifs, motif_res$motifs$motif)
selected_motif_res <- lapply(motif_res, FUN = function(x) {x[rows, ]})

create_motif_enrichment_heatmap(selected_motif_res, enrichment = "-log10(p-value)", 
                                cluster_motifs = TRUE, cluster_topics = TRUE, motif_filter = 10, horizontal = FALSE,
                                enrichment_range = c(0,100), method_cluster = "average",  font.size.motifs = 8, font.size.topics = 9)

```

Top enriched motifs
```{r motif-enrichment-ranking-plot-p0.05, fig.width=10, fig.height=7, warning=FALSE}
plots <- vector("list", ncol(motif_res$mlog10P))
names(plots) <- colnames(motif_res$mlog10P)

for( i in 1:length(plots)){
  plots[[i]] <- create_motif_enrichment_ranking_plot(motif_res, k = i, 
                                                     max.overlaps = 20, subsample = FALSE)
}

do.call(plot_grid,plots)
```


### Motif enrichment result using regions with DA p-value < 0.1

Load and compile HOMER results across topics
```{r compile-homer-motif-results-p0.1}
postfit.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Buenrostro_2018/binarized/postfit_v2"
homer.dir <- paste0(postfit.dir, "/motifanalysis-Buenrostro2018-k=10/HOMER/DA_pval_0.1_regions")
cat(sprintf("Directory of motif analysis result: %s \n", homer.dir))
homer_res_topics <- readRDS(file.path(homer.dir, "/homer_knownResults.rds"))
selected_regions <- readRDS(file.path(homer.dir, "/selected_regions.rds"))

# Compile Homer results (pvalue and ranking) across topics
motif_res <- compile_homer_motif_res(homer_res_topics)
saveRDS(motif_res, paste0(homer.dir, "/homer_motif_enrichment_results.rds"))
cat("compiled homer motif results are saved in", paste0(homer.dir, "/homer_motif_enrichment_results.rds \n"))
```

```{r all-motif-enrichment-res-p0.1, eval=FALSE, include=FALSE}
motif_table <- data.frame(motif = gsub("/.*", "", rownames(motif_res$mlog10P)),
                          round(motif_res$mlog10P,2))
DT::datatable(motif_table, rownames = F, caption = "Motif enrichment (-log10P)")
```

Top 10 motifs for each topic
```{r top-motifs-p0.1}
cat("Number of regions selected for each topic: \n")
print(mapply(nrow, selected_regions[1:(length(selected_regions)-1)]))

colnames_homer <- c("motif_name", "consensus", "P", "log10P", "Padj",  "num_target", "percent_target", "num_bg", "percent_bg")

top_motifs <- data.frame(matrix(nrow=10, ncol = length(homer_res_topics)))
colnames(top_motifs) <- names(homer_res_topics)
for (k in 1:length(homer_res_topics)){
  homer_res <- homer_res_topics[[k]]
  colnames(homer_res) <- colnames_homer
  homer_res <- homer_res %>% separate(motif_name, c("motif", "origin", "database"), "/")
  top_motifs[,k] <- head(homer_res$motif, 10)
}

DT::datatable(data.frame(rank = 1:10, top_motifs), rownames = F, caption = "Top 10 motifs enriched in each topic.")

```

Heatmap of motif enrichment -log10(p-value). 
```{r motif-heatmap-pvalue-p0.1, fig.width=7, fig.height=12}
create_motif_enrichment_heatmap(motif_res, enrichment = "-log10(p-value)", 
                                cluster_motifs = TRUE, cluster_topics = TRUE, motif_filter = 10, horizontal = FALSE,
                                enrichment_range = c(0,100), method_cluster = "average", font.size.motifs = 4, font.size.topics = 9)
```

Top enriched motifs
```{r motif-enrichment-ranking-plot-p0.1, fig.width=10, fig.height=7, warning=FALSE}
plots <- vector("list", ncol(motif_res$mlog10P))
names(plots) <- colnames(motif_res$mlog10P)

for( i in 1:length(plots)){
  plots[[i]] <- create_motif_enrichment_ranking_plot(motif_res, k = i, 
                                                     max.overlaps = 20, subsample = FALSE)
}

do.call(plot_grid,plots)
```


### Motif enrichment result using top 1% regions with largest logFC

Load and compile HOMER results across topics
```{r compile-homer-motif-results-top1percent}
postfit.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Buenrostro_2018/binarized/postfit_v2"
homer.dir <- paste0(postfit.dir, "/motifanalysis-Buenrostro2018-k=10/HOMER/DA_top1percent_regions")
cat(sprintf("Directory of motif analysis result: %s \n", homer.dir))
homer_res_topics <- readRDS(file.path(homer.dir, "/homer_knownResults.rds"))
selected_regions <- readRDS(file.path(homer.dir, "/selected_regions.rds"))

# Compile Homer results (pvalue and ranking) across topics
motif_res <- compile_homer_motif_res(homer_res_topics)
saveRDS(motif_res, paste0(homer.dir, "/homer_motif_enrichment_results.rds"))
cat("compiled homer motif results are saved in", paste0(homer.dir, "/homer_motif_enrichment_results.rds \n"))
```


```{r all-motif-enrichment-res-top1percent, eval=FALSE, include=FALSE}
motif_table <- data.frame(motif = gsub("/.*", "", rownames(motif_res$mlog10P)),
                          round(motif_res$mlog10P,2))
DT::datatable(motif_table, rownames = F, caption = "Motif enrichment (-log10P)")
```

Top 10 motifs for each topic
```{r top-motifs-top1percent}
cat("Number of regions selected for each topic: \n")
print(mapply(nrow, selected_regions[1:(length(selected_regions)-1)]))

colnames_homer <- c("motif_name", "consensus", "P", "log10P", "Padj",  "num_target", "percent_target", "num_bg", "percent_bg")

top_motifs <- data.frame(matrix(nrow=10, ncol = length(homer_res_topics)))
colnames(top_motifs) <- names(homer_res_topics)
for (k in 1:length(homer_res_topics)){
  homer_res <- homer_res_topics[[k]]
  colnames(homer_res) <- colnames_homer
  homer_res <- homer_res %>% separate(motif_name, c("motif", "origin", "database"), "/")
  top_motifs[,k] <- head(homer_res$motif, 10)
}

DT::datatable(data.frame(rank = 1:10, top_motifs), rownames = F, caption = "Top 10 motifs enriched in each topic.")

```

Heatmap of motif enrichment -log10(p-value). 
```{r motif-heatmap-pvalue-top1percent, fig.width=7, fig.height=12}
create_motif_enrichment_heatmap(motif_res, enrichment = "-log10(p-value)", 
                                cluster_motifs = TRUE, cluster_topics = TRUE, motif_filter = 10, horizontal = FALSE,
                                enrichment_range = c(0,100), method_cluster = "average", font.size.motifs = 4, font.size.topics = 9)
```

Top enriched motifs
```{r motif-enrichment-ranking-plot-top1percent, fig.width=10, fig.height=7, warning=FALSE}
plots <- vector("list", ncol(motif_res$mlog10P))
names(plots) <- colnames(motif_res$mlog10P)

for( i in 1:length(plots)){
  plots[[i]] <- create_motif_enrichment_ranking_plot(motif_res, k = i, 
                                                     max.overlaps = 20, subsample = FALSE)
}

do.call(plot_grid,plots)
```

### Motif enrichment result using top 2000 regions with largest logFC

Load and compile HOMER results across topics
```{r compile-homer-motif-results-top2000regions}
postfit.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Buenrostro_2018/binarized/postfit_v2"
homer.dir <- paste0(postfit.dir, "/motifanalysis-Buenrostro2018-k=10/HOMER/DA_top2000_regions")
cat(sprintf("Directory of motif analysis result: %s \n", homer.dir))
homer_res_topics <- readRDS(file.path(homer.dir, "/homer_knownResults.rds"))
selected_regions <- readRDS(file.path(homer.dir, "/selected_regions.rds"))

# Compile Homer results (pvalue and ranking) across topics
motif_res <- compile_homer_motif_res(homer_res_topics)
saveRDS(motif_res, paste0(homer.dir, "/homer_motif_enrichment_results.rds"))
cat("compiled homer motif results are saved in", paste0(homer.dir, "/homer_motif_enrichment_results.rds \n"))
```

```{r all-motif-enrichment-res-top2000regions, eval=FALSE, include=FALSE}
motif_table <- data.frame(motif = gsub("/.*", "", rownames(motif_res$mlog10P)),
                          round(motif_res$mlog10P,2))
DT::datatable(motif_table, rownames = F, caption = "Motif enrichment (-log10P)")
```

Top 10 motifs for each topic
```{r top-motifs-top2000regions}
cat("Number of regions selected for each topic: \n")
print(mapply(nrow, selected_regions[1:(length(selected_regions)-1)]))

colnames_homer <- c("motif_name", "consensus", "P", "log10P", "Padj",  "num_target", "percent_target", "num_bg", "percent_bg")

top_motifs <- data.frame(matrix(nrow=10, ncol = length(homer_res_topics)))
colnames(top_motifs) <- names(homer_res_topics)
for (k in 1:length(homer_res_topics)){
  homer_res <- homer_res_topics[[k]]
  colnames(homer_res) <- colnames_homer
  homer_res <- homer_res %>% separate(motif_name, c("motif", "origin", "database"), "/")
  top_motifs[,k] <- head(homer_res$motif, 10)
}

DT::datatable(data.frame(rank = 1:10, top_motifs), rownames = F, caption = "Top 10 motifs enriched in each topic.")

```

Heatmap of motif enrichment -log10(p-value). 
```{r motif-heatmap-pvalue-top2000regions, fig.width=7, fig.height=12}
create_motif_enrichment_heatmap(motif_res, enrichment = "-log10(p-value)", 
                                cluster_motifs = TRUE, cluster_topics = TRUE, motif_filter = 10, horizontal = FALSE,
                                enrichment_range = c(0,100), method_cluster = "average", font.size.motifs = 4, font.size.topics = 9)
```

Top enriched motifs
```{r motif-enrichment-ranking-plot-top2000regions, fig.width=10, fig.height=7, warning=FALSE}
plots <- vector("list", ncol(motif_res$mlog10P))
names(plots) <- colnames(motif_res$mlog10P)

for( i in 1:length(plots)){
  plots[[i]] <- create_motif_enrichment_ranking_plot(motif_res, k = i, 
                                                     max.overlaps = 20, subsample = FALSE)
}

do.call(plot_grid,plots)
```
