---
title: "Visualize and interpret topics in Lareau 2019 bone marrow scATAC-seq data"
author: "Kaixuan Luo"
output: workflowr::wflow_html
---

Here we examine and compare the topic modeling results for the 
bone marrow scATAC-seq dataset from [Lareau *et al* (2019)][Lareau-2019].

The goal of this analysis is to illustrate how the topic models
fitted to these data sets can be used to learn about structure in the
data. 

In particular, we would like to identify clusters, and interpret
clusters and topics as "cell types" or "regulatory programs".

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages used in the analysis below, as well as additional
functions that will be used to generate some of the plots.

```{r load-pkgs, message=FALSE}
library(tools)
library(dplyr)
library(fastTopics)
library(ggplot2)
library(cowplot)
source("code/plots.R")
```

Lareau 2019 bone marrow scATAC-seq dataset
------------------------------------------

### Load the data and Poisson NMF model fit
Load the data
```{r load-data}
data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Lareau_2019/bone_marrow/processed_data/"
load(file.path(data.dir, "Lareau_2019_bonemarrow.RData"))
```

About the samples:
The study used dsciATAC-seq to profile bone marrow mononuclear cells (BMMCs)
from two human donors before (untreated controls) and after stimulation, 
producing chromatin accessibility profiles for a total of 136,463 cells that passed quality filters.

Conditions: 1) Resting 2) Stimulated
```{r samples-condtion}
samples$Condition <- as.factor(samples$Condition)
table(samples$Condition)
```

15 de novo-defined clusters covering known hematopoietic cell types
```{r samples-cluster}
samples$Cluster <- as.factor(samples$Cluster)
table(samples$Cluster)

```

```{r load-PoissonNMF-fit-1}
# Load the results of running `fit_poisson_nmf` on the Lareau2019 data,
# with different algorithms, and for various choices of $k$ (the number
# of "topics").
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Lareau_2019"
load(file.path(out.dir, "/compiled.fits.Lareau2019_bonemarrow.RData"))
```

### Explore the structure of the single-cell data as inferred by the topic model.

#### Structure plot

The structure plots summarize the topic proportions in the resting (control) and stimulated conditions.

##### Resting condition
Structure plot for cells in the *resting* condition, grouped by different cell labels (15 de novo-defined clusters):

$k = 3$: 
```{r structure-plot-2, fig.width=7, fig.height=2}
fit_poisson_nmf <- fits[["fit-Lareau2019_bonemarrow-scd-ex-k=3"]]

set.seed(1)

rows <- sort(which(samples$Condition == "Resting"))
p.structure_plot <- structure_plot(select(poisson2multinom(fit_poisson_nmf), loadings = rows),
                     grouping = samples[rows,"Cluster"],
                     n = 2000,gap = 40,num_threads = 4,verbose = FALSE)
print(p.structure_plot)
```

$k = 4$: 
```{r structure-plot-3, fig.width=7, fig.height=2}
fit_poisson_nmf <- fits[["fit-Lareau2019_bonemarrow-scd-ex-k=4"]]

set.seed(1)

rows <- sort(which(samples$Condition == "Resting"))
p.structure_plot <- structure_plot(select(poisson2multinom(fit_poisson_nmf), loadings = rows),
                     grouping = samples[rows,"Cluster"],
                     n = 2000,gap = 40,num_threads = 4,verbose = FALSE)
print(p.structure_plot)
```

$k = 5$: 
```{r structure-plot-4, fig.width=7, fig.height=2}
fit_poisson_nmf <- fits[["fit-Lareau2019_bonemarrow-scd-ex-k=5"]]

set.seed(1)

rows <- sort(which(samples$Condition == "Resting"))
p.structure_plot <- structure_plot(select(poisson2multinom(fit_poisson_nmf), loadings = rows),
                     grouping = samples[rows,"Cluster"],
                     n = 2000,gap = 40,num_threads = 4,verbose = FALSE)
print(p.structure_plot)
```

$k = 6$: 
```{r structure-plot-5, fig.width=7, fig.height=2}
fit_poisson_nmf <- fits[["fit-Lareau2019_bonemarrow-scd-ex-k=6"]]

set.seed(1)

rows <- sort(which(samples$Condition == "Resting"))
p.structure_plot <- structure_plot(select(poisson2multinom(fit_poisson_nmf), loadings = rows),
                     grouping = samples[rows,"Cluster"],
                     n = 2000,gap = 40,num_threads = 4,verbose = FALSE)
print(p.structure_plot)
```

$k = 7$: 
```{r structure-plot-7, fig.width=7, fig.height=2}
fit_poisson_nmf <- fits[["fit-Lareau2019_bonemarrow-scd-ex-k=7"]]

set.seed(1)

rows <- sort(which(samples$Condition == "Resting"))
p.structure_plot <- structure_plot(select(poisson2multinom(fit_poisson_nmf), loadings = rows),
                     grouping = samples[rows,"Cluster"],
                     n = 2000,gap = 40,num_threads = 4,verbose = FALSE)
print(p.structure_plot)
```

##### Stimulated condition

Structure plot for cells in the *stimulated* condition, grouped by different cell labels:

$k = 3$: 
```{r structure-plot-8, fig.width=7, fig.height=2}
fit_poisson_nmf <- fits[["fit-Lareau2019_bonemarrow-scd-ex-k=3"]]

set.seed(1)

rows <- sort(which(samples$Condition == "Stimulated"))
p.structure_plot <- structure_plot(select(poisson2multinom(fit_poisson_nmf), loadings = rows),
                     grouping = samples[rows,"Cluster"],
                     n = 2000,gap = 40,num_threads = 4,verbose = FALSE)
print(p.structure_plot)
```

$k = 4$: 
```{r structure-plot-9, fig.width=7, fig.height=2}
fit_poisson_nmf <- fits[["fit-Lareau2019_bonemarrow-scd-ex-k=4"]]

set.seed(1)

rows <- sort(which(samples$Condition == "Stimulated"))
p.structure_plot <- structure_plot(select(poisson2multinom(fit_poisson_nmf), loadings = rows),
                     grouping = samples[rows,"Cluster"],
                     n = 2000,gap = 40,num_threads = 4,verbose = FALSE)
print(p.structure_plot)
```

$k = 5$: 
```{r structure-plot-10, fig.width=7, fig.height=2}
fit_poisson_nmf <- fits[["fit-Lareau2019_bonemarrow-scd-ex-k=5"]]

set.seed(1)

rows <- sort(which(samples$Condition == "Stimulated"))
p.structure_plot <- structure_plot(select(poisson2multinom(fit_poisson_nmf), loadings = rows),
                     grouping = samples[rows,"Cluster"],
                     n = 2000,gap = 40,num_threads = 4,verbose = FALSE)
print(p.structure_plot)
```

$k = 6$: 
```{r structure-plot-11, fig.width=7, fig.height=2}
fit_poisson_nmf <- fits[["fit-Lareau2019_bonemarrow-scd-ex-k=6"]]

set.seed(1)

rows <- sort(which(samples$Condition == "Stimulated"))
p.structure_plot <- structure_plot(select(poisson2multinom(fit_poisson_nmf), loadings = rows),
                     grouping = samples[rows,"Cluster"],
                     n = 2000,gap = 40,num_threads = 4,verbose = FALSE)
print(p.structure_plot)
```

$k = 7$: 
```{r structure-plot-12, fig.width=7, fig.height=2}
fit_poisson_nmf <- fits[["fit-Lareau2019_bonemarrow-scd-ex-k=7"]]

set.seed(1)

rows <- sort(which(samples$Condition == "Stimulated"))
p.structure_plot <- structure_plot(select(poisson2multinom(fit_poisson_nmf), loadings = rows),
                     grouping = samples[rows,"Cluster"],
                     n = 2000,gap = 40,num_threads = 4,verbose = FALSE)
print(p.structure_plot)
```


[Lareau-2019](scATACseq_data_Lareau2019.html)
