---
title: "Process data for Buenrostro 2018 scATAC-seq dataset"
author: "Kaixuan Luo"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## About Buenrostro 2018 dataset

Reference: Buenrostro, J. D. et al. Integrated Single-Cell Analysis Maps the Continuous Regulatory Landscape of Human Hematopoietic Differentiation. Cell 173, 1535–1548.e16 (2018).

Data were downloaded from GEO: [GSE96772](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE96772)

RCC directory: `/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/`

## Prepare data for topic modeling

### Using peaks called from bulk ATAC-seq samples

The Buenrostro et al paper called ATAC-seq peaks from the bulk ATAC-seq samples. 
Downloaded scATAC-seq processed data from GEO: [GSE96769](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE96769)

* Peaks called from bulk ATAC-seq samples `GSE96769_PeakFile_20160207.bed.gz`
* scATAC-seq sparse count matrix `GSE96769_scATACseq_counts.txt.gz`


```{r}
library(Matrix)
library(tools)
library(readr)
library(data.table)
```

Load the fragment counts as a 2,953 x 491,437 sparse matrix.
```{r load-data}
# The first row has the sample names
file_counts <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/GEO_data/GSE96769_scATACseq_counts.txt.gz"
sample_names <- readLines(file_counts,n = 1)
sample_names <- unlist(strsplit(sample_names,"\t",fixed = TRUE))
sample_names <- unlist(strsplit(sample_names,";",fixed = TRUE))
sample_names <- sample_names[-1]

# Load the fragment counts as sparse matrix.
dat <- fread(file_counts,sep = "\t",skip = 1)
class(dat) <- "data.frame"
names(dat) <- c("i","j","x")
counts <- sparseMatrix(i = dat$i,j = dat$j,x = dat$x)
counts <- t(counts)
rownames(counts) <- sample_names
dim(counts)
```

peaks
```{r load-peaks}
peaks <- fread("/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/GEO_data/GSE96769_PeakFile_20160207.bed.gz")
peak_names <- paste(peaks$V1,peaks$V2,peaks$V3,sep = "_")
rownames(peaks) <- peak_names
cat(sprintf("Number of peaks: %d\n",nrow(peaks)))

colnames(counts) <- peak_names

```

Plot the distribution of fragment counts.
```{r plot-distribution-1}
y <- table(summary(counts)$x)
x <- names(y)
y <- as.numeric(y)
plot(x,y,pch = 20,log = "y",cex = 0.65,
     xlab = "number of fragments mapping to peak",
     ylab = "number of peaks")
lines(x,y)

```

The supplementary Table S1 provides more details about these samples. 
https://ars.els-cdn.com/content/image/1-s2.0-S009286741830446X-mmc1.xlsx

Use the metadata.tsv from Chen et al. Genome Biology 2019 paper https://github.com/pinellolab/scATAC-benchmarking/tree/master/Real_Data/Buenrostro_2018/input/metadata.tsv)

```{r filter-samples}
samples_filtered <- read.table('/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/data/input_Chen_2019/metadata.tsv', header = TRUE, stringsAsFactors=FALSE, quote="",row.names=1)
samples_filtered$name <- rownames(samples_filtered)

idx_samples_filtered <- match(samples_filtered$name, sample_names)
sample_names_filtered <- sample_names[idx_samples_filtered]
counts <- counts[idx_samples_filtered,]
samples <- samples_filtered[,c("name", "label")]
dim(counts)
```

Remove peaks not exist in any of the cells
```{r filter-peaks}
j <- which(colSums(counts > 0) >= 1)
peaks <- peaks[j,]
counts <- counts[,j]
cat(length(j), "peaks after filtering.  \n")

```

Plot the distribution of filtered fragment counts.
```{r plot-distribution-2}
y <- table(summary(counts)$x)
x <- names(y)
y <- as.numeric(y)
plot(x,y,pch = 20,log = "y",cex = 0.65,
     xlab = "number of fragments mapping to peak",
     ylab = "number of peaks")
lines(x,y)

```

Binarize counts
```{r binarize}
binarized_counts <- as.matrix((counts > 0) + 0)
binarized_counts <- Matrix(binarized_counts, sparse = TRUE) 

dim(binarized_counts)
```

```{r save-data}
data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/processed_data/"
dir.create(data.dir, showWarnings = FALSE, recursive = TRUE)

saveRDS(counts, file.path(data.dir, "counts_Buenrostro_2018.rds"))
saveRDS(binarized_counts, file.path(data.dir, "binarized_counts_Buenrostro_2018.rds"))

save(list = c("samples", "peaks", "counts"), 
     file = file.path(data.dir, "Buenrostro_2018.RData"))

counts <- binarized_counts
save(list = c("samples", "peaks", "counts"), 
     file = file.path(data.dir, "Buenrostro_2018_binarized.RData"))

```

```{r}
data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/processed_data/"
load(file.path(data.dir, "Buenrostro_2018.RData"))
cat(sprintf("Loaded %d x %d counts matrix.\n",nrow(counts),ncol(counts)))

cat(sprintf("Number of samples (cells): %d\n",nrow(counts)))
cat(sprintf("Number of peaks: %d\n",ncol(counts)))
cat(sprintf("Proportion of counts that are non-zero: %0.1f%%.\n",
            100*mean(counts > 0)))

```

### Using peaks called from aggregated single-cell ATAC-seq data processed by Chen et al. Genome Biology 2019
* Downloaded processed peaks and BAM files from the scATAC-benchmarking website (Chen et al. Genome Biology 2019) (https://github.com/pinellolab/scATAC-benchmarking/)
* Count paired-end scATAC-seq reads in peaks using `chromVAR`, following the pipeline at https://nbviewer.jupyter.org/github/pinellolab/scATAC-benchmarking/blob/master/Real_Data/Buenrostro_2018/run_methods/chromVAR/chromVAR_buenrostro2018_motifs.ipynb?flush_cache=true

Load chromVAR and related packages
```{r}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("chromVAR")
# BiocManager::install("motifmatchr")
# BiocManager::install("BSgenome.Hsapiens.UCSC.hg19")
# BiocManager::install("JASPAR2016")
```

```{r}
library(chromVAR)
library(motifmatchr)
library(Matrix)
library(SummarizedExperiment)
library(BiocParallel)
library('JASPAR2016')
library(BSgenome.Hsapiens.UCSC.hg19)

register(MulticoreParam(10))

data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/processed_data_Chen2019pipeline/chromVAR/"
dir.create(data.dir, showWarnings = FALSE, recursive = TRUE)
```

```{r}
samples <- read.table('/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/data/input_Chen_2019/metadata.tsv', header = TRUE, stringsAsFactors=FALSE, quote="",row.names=1)
```

```{r}
peakfile <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/data/input_Chen_2019/combined.sorted.merged.bed"
peaks <- getPeaks(peakfile, sort_peaks = TRUE)
peaks <- resize(peaks, width = 500, fix = "center")

cat(length(peaks), "peaks \n")
```

```{r}
dir_bamfiles <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/data/input_Chen_2019/sc-bams_nodup/"
bamfiles <- list.files(path = dir_bamfiles, pattern = "\\.bam$")
cellnames <- sapply(strsplit(bamfiles,'.',fixed = TRUE), "[[", 1)
sum(cellnames == rownames(samples))
```

count scATAC-seq paired-end fragments in the peaks
```{r, eval=FALSE}
fragment_counts <- getCounts(file.path(dir_bamfiles,bamfile), 
                             peaks, 
                             paired = TRUE, 
                             by_rg = TRUE, 
                             format = "bam", 
                             colData = data.frame(celltype = cellnames))

saveRDS(fragment_counts, file.path(data.dir, "fragment_counts_scPeaks_chromVAR_Buenrostro_2018.rds"))

```

```{r}
fragment_counts <- readRDS(file.path(data.dir, "fragment_counts_scPeaks_chromVAR_Buenrostro_2018.rds"))
fragment_counts

```

```{r}
counts <- assay(fragment_counts)
counts <- t(counts)

peaks <- as.data.frame(peaks)[,1:3]
colnames(peaks) <- c("chr", "start", "end")
peak_names <- paste(peaks$chr, peaks$start, peaks$end, sep = "_")
colnames(counts) <- peak_names

```

Filter peaks using `filterPeaks` in `chromVAR`
```{r}
idx_peaks_filtered <- filterPeaks(fragment_counts, min_fragments_per_peak = 1, non_overlapping = TRUE, ix_return = TRUE)
peaks <- peaks[idx_peaks_filtered,]
counts <- counts[,idx_peaks_filtered]
cat(length(idx_peaks_filtered), "peaks after filtering.  \n")

```

Plot the distribution of filtered fragment counts.
```{r}
y <- table(summary(counts)$x)
x <- names(y)
y <- as.numeric(y)
plot(x,y,pch = 20,log = "y",cex = 0.65,
     xlab = "number of fragments mapping to peak",
     ylab = "number of peaks")
lines(x,y)

```

Binarize counts
```{r}
binarized_counts <- as.matrix((counts > 0) + 0)
binarized_counts <- Matrix(binarized_counts, sparse = TRUE) 
dim(binarized_counts)
```

```{r}
saveRDS(counts, file.path(data.dir, "counts_scPeaks_Buenrostro_2018.rds"))
saveRDS(binarized_counts, file.path(data.dir, "binarized_counts_scPeaks_Buenrostro_2018.rds"))

save(list = c("samples", "peaks", "counts"), 
     file = file.path(data.dir, "Buenrostro_2018_scPeaks.RData"))

counts <- binarized_counts
save(list = c("samples", "peaks", "counts"), 
     file = file.path(data.dir, "Buenrostro_2018_binarized_scPeaks.RData"))

```

```{r}
data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/processed_data_Chen2019pipeline/chromVAR/"

load(file.path(data.dir, "Buenrostro_2018_binarized_scPeaks.RData"))
cat(sprintf("Loaded %d x %d counts matrix.\n",nrow(counts),ncol(counts)))

cat(sprintf("Number of samples (cells): %d\n",nrow(counts)))
cat(sprintf("Number of peaks: %d\n",ncol(counts)))
cat(sprintf("Proportion of counts that are non-zero: %0.1f%%.\n",
            100*mean(counts > 0)))

```

