---
title: "Process data for Buenrostro 2018 scATAC-seq dataset"
author: "Kaixuan Luo"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## About Buenrostro 2018 dataset

Reference: Buenrostro, J. D. et al. Integrated Single-Cell Analysis Maps the Continuous Regulatory Landscape of Human Hematopoietic Differentiation. Cell 173, 1535–1548.e16 (2018).

Data were downloaded from GEO: [GSE96772](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE96772)

RCC directory: `/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/`

## Processed data

Downloaded scATAC-seq processed data from GEO: [GSE96769](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE96769)

* `GSE96769_PeakFile_20160207.bed.gz`
* `GSE96769_scATACseq_counts.txt.gz`

Downloaded processed data from the scATAC-benchmarking website (Chen et al. Genome Biology 2019) (https://github.com/pinellolab/scATAC-benchmarking/)

## Prepare data for topic modeling
```{r}
library(Matrix)
library(tools)
library(readr)
library(data.table)

```

count scATAC-seq reads in peaks
```{bash, eval=FALSE}
sbatch ~/projects/scATACseq-topics/scripts/count_reads_peaks_Buenrostro_2018.sbatch
```

combine counts from all samples
```{r}
dir_readcount_output <- '/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/data/count_reads_peaks_output/'
files <- list.files(dir_readcount_output, pattern = "\\.txt$")
sample_names <- sapply(strsplit(files,'\\.'),'[', 1)
cat(length(sample_names), "samples. \n")
```

```{r, eval=FALSE}
datalist <- lapply(files, function(x)fread(file.path(dir_readcount_output,x))$V4)
counts <- do.call("cbind", datalist)
```

```{r}
data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/processed_data_Chen2019pipeline/"
dir.create(data.dir, showWarnings = FALSE, recursive = TRUE)

# saveRDS(counts, file.path(data.dir, "raw_counts_Buenrostro_2018.rds"))
counts <- readRDS(file.path(data.dir, "raw_counts_Buenrostro_2018.rds"))
```

Processing count files (adapted from Chen et al. Genome Biology 2019 paper https://github.com/pinellolab/scATAC-benchmarking/blob/master/Real_Data/Buenrostro_2018/run_methods/Control/Control_buenrostro2018.ipynb)

```{r}
peaks <- read.csv("/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/data/input_Chen_2019/combined.sorted.merged.bed",
                      sep = '\t',header=FALSE,stringsAsFactors=FALSE)
peaknames <- paste(peaks$V1,peaks$V2,peaks$V3,sep = "_")
rownames(peaks) <- peaknames
cat(sprintf("Number of peaks: %d\n",nrow(peaks)))

head(peaknames)

colnames(counts) <- sample_names
rownames(counts) <- peaknames

dim(counts)
```

Sample labels
```{r}
samples <- read.table('/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/data/input_Chen_2019/metadata.tsv', header = TRUE, stringsAsFactors=FALSE, quote="",row.names=1)

head(samples)
```

Filter peaks (filter out peaks with counts in < 1% samples)
```{r}
# adapted from Chen et al. Genome Biology 2019 paper https://github.com/pinellolab/scATAC-benchmarking/blob/master/Real_Data/Buenrostro_2018/run_methods/Control/Control_buenrostro2018.ipynb
# filter peaks with counts in at least 1% samples
filter_peaks <- function (datafr,cutoff = 0.01){
    binary_mat = as.matrix((datafr > 0) + 0)
    binary_mat = Matrix(binary_mat, sparse = TRUE) 
    num_cells_ncounted = Matrix::rowSums(binary_mat)
    ncounts = binary_mat[num_cells_ncounted >= dim(binary_mat)[2]*cutoff,]
    ncounts = ncounts[rowSums(ncounts) > 0,] 
    
    options(repr.plot.width=4, repr.plot.height=4)
    hist(log10(num_cells_ncounted),main="No. of Cells Each Site is Observed In",breaks=50)
    abline(v=log10(min(num_cells_ncounted[num_cells_ncounted >= dim(binary_mat)[2]*cutoff])),lwd=2,col="indianred")
#     hist(log10(new_counts),main="Number of Sites Each Cell Uses",breaks=50)
    peaks_selected = rownames(ncounts)
    return(peaks_selected)
}
```

```{r}
peaks_selected <- filter_peaks(counts)
counts <- counts[peaks_selected,]
peaks <- peaks[peaks_selected, ]

counts <- t(counts)
counts <- Matrix(counts, sparse = TRUE) 
dim(counts)

```

Binarize counts
```{r}
binarized_counts <- as.matrix((counts > 0) + 0)
binarized_counts <- Matrix(binarized_counts, sparse = TRUE) 

dim(binarized_counts)
```

```{r, eval=FALSE, include=TRUE}
saveRDS(counts, file.path(data.dir, "counts_Buenrostro_2018.rds"))
saveRDS(binarized_counts, file.path(data.dir, "binarized_counts_Buenrostro_2018.rds"))

save(list = c("samples", "peaks", "counts"), 
     file = file.path(data.dir, "Buenrostro_2018_counts.RData"))

counts <- binarized_counts
save(list = c("samples", "peaks", "counts"), 
     file = file.path(data.dir, "Buenrostro_2018_binarized_counts.RData"))

```

```{r}
data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/processed_data/"
load(file.path(data.dir, "Buenrostro_2018_counts.RData"))
cat(sprintf("Loaded %d x %d counts matrix.\n",nrow(counts),ncol(counts)))

cat(sprintf("Number of samples (cells): %d\n",nrow(counts)))
cat(sprintf("Number of peaks: %d\n",ncol(counts)))
cat(sprintf("Proportion of counts that are non-zero: %0.1f%%.\n",
            100*mean(counts > 0)))

```


