---
title: "Selected figures for Buenrostro *et al* (2018) scATAC-seq results"
author: Kaixuan Luo
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: false
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = FALSE,results = "hold",
                      fig.align = "center", dpi = 600)
```

## Load packages and some functions used in this analysis
```{r load-pkgs, message=FALSE}
library(Matrix)
library(fastTopics)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(cowplot)
library(plotly)
library(htmlwidgets)
library(DT)
library(reshape2)
library(Logolas)
library(grid)
source("code/functions_for_assessing_fits.R")
source("code/motif_analysis.R")
source("code/plots.R")
```

## Load data and topic model results
## Load data and topic model results
Load the **binarized** data and the $k = 11$ Poisson NMF fit results 
```{r load-data}
data.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/data/Buenrostro_2018/processed_data_Chen2019pipeline/"
load(file.path(data.dir, "Buenrostro_2018_binarized_counts.RData"))
cat(sprintf("%d x %d counts matrix.\n",nrow(counts),ncol(counts)))
```

## Progress plot showing convergence of the optimization methods

Load the results of running `fit_poisson_nmf` on the Cusanovich_2018 data,
with different algorithms, and for various choices of $k$ (the number
of "topics").

```{r load-compiled-fit}
out.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Buenrostro_2018_Chen2019pipeline/binarized/"
load(file.path(out.dir, "/compiled.fits.Buenrostro2018-binarized.RData"))
```

## Structure plot grouped by tissues

```{r load-fit}
fit.dir <- "/project2/mstephens/kevinluo/scATACseq-topics/output/Buenrostro_2018_Chen2019pipeline/binarized/"
fit <- readRDS(file.path(fit.dir, "/fit-Buenrostro2018-binarized-scd-ex-k=11.rds"))$fit
fit <- poisson2multinom(fit)
```

## Visualize by Structure plot grouped by cell labels.
```{r structure-plot, fig.width=7, fig.height=2.5, message=FALSE}
set.seed(10)
colors_topics <- c("#a6cee3","#1f78b4","#b2df8a","#33a02c","#fb9a99","#e31a1c",
                   "#fdbf6f","#ff7f00","#cab2d6","#6a3d9a","#ffff99","#b15928",
                   "gray")
samples$label <- factor(samples$label, levels = c("HSC", "MPP", "CMP", "GMP", "mono", "MEP", "LMPP", "CLP", "pDC", "UNK"))

p.structure <- structure_plot(fit,
                     grouping = samples[, "label"],n = Inf,gap = 40,
                     perplexity = 50,topics = 1:11,colors = colors_topics,
                     num_threads = 4,verbose = FALSE)

print(p.structure)

ggsave("paper/Buenrostro2018_structure_plot.eps", width = 7, height = 2.5, dpi = 600)

```

## TF motif enrichment

```{r compile-homer-motif-results}
# Compile Homer results across topics
homer.dir <- paste0(out.dir, "/postfit_v2/motifanalysis-Buenrostro2018-k=11-quantile/HOMER/")
cat(sprintf("Directory of motif analysis result: %s \n", homer.dir))
homer_res_topics <- readRDS(file.path(homer.dir, "/homer_knownResults.rds"))
selected_regions <- readRDS(file.path(homer.dir, "/selected_regions.rds"))

# Compile Homer results (pvalue and ranking) across topics
motif_res <- compile_homer_motif_res(homer_res_topics)
# saveRDS(motif_res, paste0(homer.dir, "/homer_motif_enrichment_results.rds"))
# motif_res <- readRDS(paste0(homer.dir, "/homer_motif_enrichment_results.rds"))

```


Plot enrichment (-log10 p-value) vs. ranking of the motifs

```{r motif-enrichment-ranking-plot, fig.width=10, fig.height=7, warning=FALSE}

plots <- vector("list", ncol(motif_res$mlog10P))
names(plots) <- colnames(motif_res$mlog10P)

for( i in 1:length(plots)){
  plots[[i]] <- create_motif_enrichment_ranking_plot(motif_res, k = i, point.size = 1,
                                                     max.overlaps = 20, subsample = FALSE)
}

do.call(plot_grid,plots)

ggsave("paper/Buenrostro2018_motif_enrichment_ranking_topics.eps", width = 10, height = 7, dpi = 600)

```

**Topic 1 and topic 4 example**
```{r motif-enrichment-ranking-plot-2, fig.width=5, fig.height=4, warning=FALSE}

create_motif_enrichment_ranking_plot(motif_res, k = 1, title = "TF motif enrichment in topic 1", point.size = 1, max.overlaps = 20, subsample = FALSE)
ggsave("paper/Buenrostro2018_motif_enrichment_ranking_topic1.eps", width = 5, height = 4, dpi = 600)


create_motif_enrichment_ranking_plot(motif_res, k = 4, title = "TF motif enrichment in topic 4", point.size = 1, max.overlaps = 20, subsample = FALSE)
ggsave("paper/Buenrostro2018_motif_enrichment_ranking_topic4.eps", width = 5, height = 4, dpi = 600)

```
